---
title: "BC_demography_02_agedistribution"
author: "Martin Monkman"
date: "February 14, 2016"
output: html_document
---

## Plot the age distribution in British Columbia, 1971-2015

### data

Source:
Statistics Canada, CANSIM 
[Table 051-0001: Estimates of population, by age group and sex for July 1, Canada, provinces and territories annual](http://www5.statcan.gc.ca/cansim/a26?lang=eng&retrLang=eng&id=0510001&&pattern=&stByVal=1&p1=1&p2=37&tabMode=dataTable&csid=)

The source file contains population for Canada, the provinces and territories, by age (and age group), and by sex, for the years 1976-2016 (as of this writing).

Source file:  (./data_source/cansim-0510001-eng-8333626124709750278.csv)


The CANSIM extract requires some carpentry to be useful for the purposes of this plotting. The tidied version becomes a data frame `data_agedistrib01`.

**Notes:** 

1. The table originally tabulated the last category as "90 years and over"; starting with 2001 this category was expanded to single years 90 through 99, and a "100 years and over" category was added. Depending on the year(s) of interest, the treatment of the extreme category will vary. 

2. The extreme age categories appear for all years, but are empty values. These lead to "parsing failures" when reading the source .csv file.


```{r, echo=FALSE, message=FALSE}

# load the necessary packages
# data manipulation
library(readr)
library(tidyr)
library(dplyr)
# graphics
library(ggplot2)
library(scales)
# tables
library(xtable)
# install necessary packages
# utilities
library(zoo)


# PLOTTING ESSENTIALS

# create a chart theme & colour palette for use in ggplot2
# this theme is the basic background for a BC Stats chart
#
theme_BCStats <- 
  theme_bw() +
  theme(
    panel.border = element_rect(colour="white"),
    axis.line = element_line(colour="black"),
    legend.position=c(1,0), 
    legend.justification=c(1,0),
    legend.title = element_text(size=12),
    legend.text = element_text(size=11),
    axis.title = element_text(size=16),
    axis.text = element_text(size=16),
    plot.title = element_text(size=18)
  )
#

# colour palette for BC Stats charts (use with scale_colour_manual)
palette_BCStats <- c("#234275", "#E3A82B", "#26BDEF", "#11CC33", "#D3E2ED", "8A8A8A")

# grayscale for fill (use with scale_fill_manual)
palette_BCStats_fill <- c("#3F3F3F", "#ABABAB", "#DFDFDF", "#969696", "#838383", "8A8A8A")



```



```{r, echo=FALSE, message=FALSE}

# read the data
setwd("C:/@data/GitHub/BC_demography")
#

data_agedistrib <- read_csv("./data_source/00510001-eng.csv")

ls.str(data_agedistrib01)

unique(data_agedistrib$AGE)
unique(data_agedistrib$GEO)
unique(data_agedistrib$Ref_Date)
unique(data_agedistrib$SEX)

# -------------------------------------

# filter for British Columbia, both sexes, 2016 distribution

data_agedistrib01 <- data_agedistrib %>%
  filter(GEO == "British Columbia") %>%
  filter(Ref_Date == "2016") %>%
  filter(SEX == "Both sexes")

unique(data_agedistrib01$AGE)
unique(data_agedistrib01$GEO)
unique(data_agedistrib01$Ref_Date)
unique(data_agedistrib01$SEX)


drop.age.vals <- c("All ages", "0 to 4 years", "5 to 9 years",
                   "10 to 14 years", "15 to 19 years", 
                   "20 to 24 years", "25 to 29 years", 
                   "30 to 34 years", "35 to 39 years", 
                   "40 to 44 years", "45 to 49 years", 
                   "50 to 54 years", "55 to 59 years", 
                   "60 to 64 years", "65 to 69 years", 
                   "70 to 74 years", "75 to 79 years", 
                   "80 to 84 years", "85 to 89 years", 
                   "90 to 94 years", "95 to 99 years",
                   "0 to 14 years", "0 to 15 years", "0 to 16 years", "0 to 17 years",
                   "15 to 49 years", "15 to 64 years", "16 to 64 years", "17 to 64 years",
                   "18 years and over", "18 to 64 years", "18 to 24 years", "25 to 44 years",
                   "45 to 64 years", "65 years and over", "90 years and over", 
                   "Median age (years)")

# filter out the values
data_agedistrib01 <- data_agedistrib01 %>%
  filter(!AGE %in% drop.age.vals)

# check the list again
unique(data_agedistrib01$AGE)


# create a new variable age.num that converts the age variable to an integer 
# (instead of a string that has " year" AND no padding, so an alpha sort will yield inappropriate results)
# step 1: use gsub to replace " years" (and its variants) with blank in string
# step 2: set as.integer
data_agedistrib01$age.num <- gsub(" years", "", data_agedistrib01$AGE)
data_agedistrib01$age.num <- gsub(" year", "", data_agedistrib01$age.num)
data_agedistrib01$age.num <- as.integer(gsub(" and over", "", data_agedistrib01$age.num))

unique(data_agedistrib01$age.num)


```

### Basic plot

```{r}

# basic
ggplot(data_agedistrib01, aes(x=age.num, y=Value)) +
  geom_bar(stat="identity", fill="#234275")



```





-30-