---
title: "BC_demography_02_agedistribution"
author: "Martin Monkman"
date: "February 14, 2016"
output: html_document
---

## Plot the age distribution in British Columbia, 1971-2016

### data

Source:
Statistics Canada, CANSIM 
[Table 051-0001: Estimates of population, by age group and sex for July 1, Canada, provinces and territories annual](http://www5.statcan.gc.ca/cansim/a26?lang=eng&retrLang=eng&id=0510001&&pattern=&stByVal=1&p1=1&p2=37&tabMode=dataTable&csid=)

The source file contains population for Canada, the provinces and territories, by age (and age group), and by sex, for the years 1976-2016 (as of this writing).

Source file:  (./data_source/cansim-0510001-eng-8333626124709750278.csv)


The CANSIM extract requires some carpentry to be useful for the purposes of this plotting. The tidied version becomes a data frame `data_agedistrib_BC_2016`.

**Notes:** 

1. The table originally tabulated the last category as "90 years and over"; starting with 2001 this category was expanded to single years 90 through 99, and a "100 years and over" category was added. Depending on the year(s) of interest, the treatment of the extreme category will vary. 

2. The extreme age categories appear for all years, but are empty values. These lead to "parsing failures" when reading the source .csv file.


```{r, echo=FALSE, message=FALSE}

# load the necessary packages
# data manipulation
library(readr)
library(tidyr)
library(dplyr)
# graphics
library(ggplot2)
library(scales)
# tables
library(xtable)
# install necessary packages
# utilities
library(zoo)


# PLOTTING ESSENTIALS

# create a chart theme & colour palette for use in ggplot2
# this theme is the basic background for a BC Stats chart
#
theme_BCStats <- 
  theme_bw() +
  theme(
    panel.border = element_rect(colour="white"),
    axis.line = element_line(colour="black"),
    legend.position=c(1,0), 
    legend.justification=c(1,0),
    legend.title = element_text(size=12),
    legend.text = element_text(size=11),
    axis.title = element_text(size=16),
    axis.text = element_text(size=16),
    plot.title = element_text(size=18)
  )
#

# colour palette for BC Stats charts (use with scale_colour_manual)
palette_BCStats <- c("#234275", "#E3A82B", "#26BDEF", "#11CC33", "#D3E2ED", "8A8A8A")

# grayscale for fill (use with scale_fill_manual)
palette_BCStats_fill <- c("#3F3F3F", "#ABABAB", "#DFDFDF", "#969696", "#838383", "8A8A8A")



```



```{r, echo=FALSE, message=FALSE}

# read the data
setwd("C:/@data/GitHub/BC-demography")
#

data_agedistrib <- read_csv("./data_source/00510001-eng.csv")

ls.str(data_agedistrib)

unique(data_agedistrib$AGE)
unique(data_agedistrib$GEO)
unique(data_agedistrib$Ref_Date)
unique(data_agedistrib$SEX)

# -------------------------------------

# filter for British Columbia, both sexes, 2016 distribution

data_agedistrib_BC_2016 <- data_agedistrib %>%
  select(GEO, Ref_Date, SEX, AGE,  Value) %>%
  filter(GEO == "British Columbia") %>%
  filter(Ref_Date == "2016") %>%
  filter(SEX == "Both sexes")

ls.str(data_agedistrib_BC_2016)

unique(data_agedistrib_BC_2016$AGE)
unique(data_agedistrib_BC_2016$GEO)
unique(data_agedistrib_BC_2016$Ref_Date)
unique(data_agedistrib_BC_2016$SEX)


drop.age.vals_2 <- c("All ages", "0 to 4 years", "5 to 9 years",
                   "10 to 14 years", "15 to 19 years", 
                   "20 to 24 years", "25 to 29 years", 
                   "30 to 34 years", "35 to 39 years", 
                   "40 to 44 years", "45 to 49 years", 
                   "50 to 54 years", "55 to 59 years", 
                   "60 to 64 years", "65 to 69 years", 
                   "70 to 74 years", "75 to 79 years", 
                   "80 to 84 years", "85 to 89 years", 
                   "90 to 94 years", "95 to 99 years",
                   "0 to 14 years", "0 to 15 years", "0 to 16 years", "0 to 17 years",
                   "15 to 49 years", "15 to 64 years", "16 to 64 years", "17 to 64 years",
                   "18 years and over", "18 to 64 years", "18 to 24 years", "25 to 44 years",
                   "45 to 64 years", "65 years and over", "90 years and over", 
                   "Median age (years)")

# filter out the values
data_agedistrib_BC_2016 <- data_agedistrib_BC_2016 %>%
  filter(!AGE %in% drop.age.vals_2)

# check the list again
unique(data_agedistrib_BC_2016$AGE)


# create a new variable age.num that converts the age variable to an integer 
# (instead of a string that has " year" AND no padding, so an alpha sort will yield inappropriate results)
# step 1: use gsub to replace " years" (and its variants) with blank in string
# step 2: set as.integer
data_agedistrib_BC_2016$age.num <- gsub(" years", "", data_agedistrib_BC_2016$AGE)
data_agedistrib_BC_2016$age.num <- gsub(" year", "", data_agedistrib_BC_2016$age.num)
data_agedistrib_BC_2016$age.num <- as.integer(gsub(" and over", "", data_agedistrib_BC_2016$age.num))

unique(data_agedistrib_BC_2016$age.num)

# calculate percentage of total population for each age group
data_agedistrib_BC_2016 <- data_agedistrib_BC_2016 %>%
  mutate(year_pct = (Value / sum(Value) * 100)) 


```

### Plot: age distribution, British Columbia 2016

This set of charts shows the distribution of the B.C. population, by one year age groups.

The first group shows the absolute number of people, and the second shows the percentage of the population in each group (useful for comparisons across time).


The first version of the plot is a basic version. The second version adds an assortment of formatting:

* axis length and grid specifications
* axis titles
* chart title, subtitle, and source

Note that the X axis scale is such that it follows the five year increments of the Census of Canada.



```{r}

# basic
plot_agedist_BC_2016 <- ggplot(data_agedistrib_BC_2016, aes(x=age.num, y=Value)) +
  geom_bar(stat="identity", fill="#234275")

plot_agedist_BC_2016


# Add formatting

plot_agedist_BC_2016 <- plot_agedist_BC_2016 +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  labs(title = "British Columbia, Age distribution, 2016",
       y = "population", x = NULL,
       caption ="Source: Statistics Canada, CANSIM Table 051-0001") +
  scale_x_continuous(breaks = seq(0, 100, 10)) +
#  scale_x_discrete(breaks = c("10", "20", "30", "40", "50", 
#                              "60", "70", "80", "90", "100+")) +
  scale_y_continuous(limits = c(0, 80000), breaks = seq(0, 80000, by = 10000),
                     labels = comma)

plot_agedist_BC_2016

# save the plot as a png file
png("plot_agedist_BC_2016.png", width=1024, height=768, res=120)
plot(plot_agedist_BC_2016)
dev.off()
  
```


Repeat, but with percentage

```{r}

# basic
plot_agedist_BC_2016_pct <- ggplot(data_agedistrib_BC_2016, aes(x=age.num, y=year_pct)) +
  geom_bar(stat="identity", fill="#234275")

plot_agedist_BC_2016_pct


# Add formatting

plot_agedist_BC_2016_pct <- plot_agedist_BC_2016_pct +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  labs(title = "British Columbia, Age distribution, 2016",
       y = "percent of population", x = NULL,
       caption ="Source: Statistics Canada, CANSIM Table 051-0001") +
  scale_x_continuous(breaks = seq(0, 100, 10)) +
  scale_y_continuous(limits = c(0, 2.5), breaks = seq(0, 2.5, by = 0.5)) 

plot_agedist_BC_2016_pct


# save the plot as a png file
png("plot_agedist_BC_2016_pct.png", width=1024, height=768, res=120)
plot(plot_agedist_BC_2016_pct)
dev.off()
  
```


```{r}

# highlight the boomers

plot_agedist_BC_2016_shade <- plot_agedist_BC_2016 +
  # add shaded rectangle for the projection range
  annotate("rect", xmin=50, xmax=70, ymin=0, ymax=80000, alpha=0.1, fill="blue") +
  annotate("text", x=60, y=77000,  label="Boomers")

plot_agedist_BC_2016_shade


```


The total population of British Columbia has increased from 2.2 million in 1971 to 4.7 million in 2015. The population surpassed 3 million in 1986, and 4 million in 1999.

#### Density Plot

An effective way to compare the distribution of the population.

First, create a cleaned up version of the data for 1971.

Note that the age categories are different than those used for 2016!

```{r}
# filter for British Columbia, both sexes, 1971 distribution

data_agedistrib_BC_1971 <- data_agedistrib %>%
  select(GEO, Ref_Date, SEX, AGE,  Value) %>%
  filter(GEO == "British Columbia") %>%
  filter(Ref_Date == "1971") %>%
  filter(SEX == "Both sexes")

ls.str(data_agedistrib_BC_1971)

unique(data_agedistrib_BC_1971$AGE)
unique(data_agedistrib_BC_1971$GEO)
unique(data_agedistrib_BC_1971$Ref_Date)
unique(data_agedistrib_BC_1971$SEX)


drop.age.vals_1 <- c("All ages", "0 to 4 years", "5 to 9 years",
                   "10 to 14 years", "15 to 19 years", 
                   "20 to 24 years", "25 to 29 years", 
                   "30 to 34 years", "35 to 39 years", 
                   "40 to 44 years", "45 to 49 years", 
                   "50 to 54 years", "55 to 59 years", 
                   "60 to 64 years", "65 to 69 years", 
                   "70 to 74 years", "75 to 79 years", 
                   "80 to 84 years", "85 to 89 years", 
                   "90 to 94 years", "95 to 99 years",
                   "0 to 14 years", "0 to 15 years", "0 to 16 years", "0 to 17 years",
                   "15 to 49 years", "15 to 64 years", "16 to 64 years", "17 to 64 years",
                   "18 years and over", "18 to 64 years", "18 to 24 years", "25 to 44 years",
                   "45 to 64 years", "65 years and over", 
#                   "90 years and over", 
#                   "100 years and over", 
                   "Median age (years)")

drop.age.vals_1

# filter out the values
data_agedistrib_BC_1971 <- data_agedistrib_BC_1971 %>%
  filter(!AGE %in% drop.age.vals_1)

# check the list again
unique(data_agedistrib_BC_1971$AGE)


# create a new variable age.num that converts the age variable to an integer 
# (instead of a string that has " year" AND no padding, so an alpha sort will yield inappropriate results)
# step 1: use gsub to replace " years" (and its variants) with blank in string
# step 2: set as.integer
data_agedistrib_BC_1971$age.num <- gsub(" years", "", data_agedistrib_BC_1971$AGE)
data_agedistrib_BC_1971$age.num <- gsub(" year", "", data_agedistrib_BC_1971$age.num)
data_agedistrib_BC_1971$age.num <- as.integer(gsub(" and over", "", data_agedistrib_BC_1971$age.num))

unique(data_agedistrib_BC_1971$age.num)

# calculate percentage of total population for each age group
data_agedistrib_BC_1971 <- data_agedistrib_BC_1971 %>%
  mutate(year_pct = (Value / sum(Value) * 100)) 


```

Merge 1971 data frame with 2016

```{r}

data_agedistrib_BC_19712016 <- rbind(data_agedistrib_BC_1971, data_agedistrib_BC_2016)

ls.str(data_agedistrib_BC_19712016)
unique(data_agedistrib_BC_19712016$age.num)

```


Plot

```{r}

plot_agedistrib_BC_19712016 <- 
  ggplot(data_agedistrib_BC_19712016, aes(x=age.num, fill=Ref_Date)) +
  geom_density(alpha = 0.25)

plot_agedistrib_BC_19712016


ggplot(data_agedistrib_BC_19712016, aes(x=age.num, fill=Ref_Date)) +
  geom_density()



```



-30-