---
title: "BC_demography_01_medianage"
author: "Martin Monkman"
date: "February 8, 2016"
output: html_document
---

## Plot of the change in median age in British Columbia, 1971-2015

### data

Source file:  (data_source/cansim-0510001-eng-7291498208024636066.csv)

The CANSIM extract was cleaned up to produce a wide table that contained only the year (as the first variable) and the name of the region (Canada, province, or territory). 

Note that the Northwest Territories splits in 1991, into Northwest Territories and Nunavut.

The table is then melted into long format.


```{r, echo=FALSE, message=FALSE}

# load the necessary packages
library(tidyr)
library(dplyr)

# read the data
#setwd("~/GitHub/BC_demography")
data_medianage <- read.csv("./data_source/data_medianage.csv", header = TRUE)
#data_medianage <- read.csv("data_medianage.csv", header = TRUE)

# turn the wide version of data_medianage into long format
data_medianage_long <- data_medianage %>%
  gather(region, median_age, -year)

# filter the long frame to include only B.C. 
data_median_age_BC <- filter(data_medianage_long, region == "British.Columbia")

```


```{r, echo=FALSE, message=FALSE}

# PLOTTING ESSENTIALS

#load the necessary packages for plotting
library(ggplot2)

# -----

# create a chart theme & colour palette for use in ggplot2
# this theme is the basic background for a BC Stats chart
#
theme_BCStats <- 
  theme_bw() +
  theme(
    panel.border = element_rect(colour="white"),
    axis.line = element_line(colour="black"),
    legend.position=c(1,0), 
    legend.justification=c(1,0),
    legend.title = element_text(size=12),
    legend.text = element_text(size=11),
    axis.title = element_text(size=12),
    axis.text = element_text(size=12)
  )
#

# colour palette for BC Stats charts

palette_BCStats <- c("#234275", "#E3A82B", "#26BDEF", "#11CC33", "#D3E2ED", "8A8A8A")


```


### B.C. median age plot

The first version of the plot is a basic version of the plot, with only the line weight and colour specified.

The second version adds an assortment of formatting:

* axis length and grid specifications
* axis titles
* chart title

Note that the X axis scale is such that it follows the five year increments of the Census of Canada.





```{r, echo=FALSE, message=FALSE}


# basic
plot_median_age_BC <- 
  ggplot(data_median_age_BC, aes(x=year, y=median_age, colour=region)) +
  geom_line(size = 1.5)

plot_median_age_BC

# fix the formatting

plot_median_age_BC +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  ggtitle("Median age, British Columbia, 1971-2015") +
  ylab("median age") +
  scale_x_continuous(breaks = seq(1971, 2016, by = 5)) +
  scale_y_continuous(limits = c(20, 45), breaks = seq(20, 45, by = 5))


```

### B.C. and Canada median age plot

This version adds a second line to allow for the comparison of the change in the median age of British Columbians to that of Canadians overall.


```{r, echo=FALSE, message=FALSE}

# filter the long frame to include only B.C. and Canada
regions <- c("British.Columbia", "Canada")
data_median_age_BC_Can <- 
  filter(data_medianage_long, region %in% regions)

# basic
plot_median_age_BC_Can <- 
  ggplot(data_median_age_BC_Can, aes(x=year, y=median_age, colour = region)) +
  geom_line(size = 1.5)

# fix the formatting
plot_median_age_BC_Can +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  ggtitle("Median age, British Columbia and Canada, 1971-2015") +
  ylab("median age") +
  scale_x_continuous(breaks = seq(1971, 2016, by = 5)) +
  scale_y_continuous(limits = c(20, 45), breaks = seq(20, 45, by = 5))


```

### Conventional wisdom: B.C. is an old province

The above chart makes it plain that British Columbians are indeed older than Canadians as a whole. 

But what may not be known is that British Columbia, while once the oldest province, no longer has that distinction. 

Between 1994 and 2001, Quebec had the highest median age. From 2002 through 2004, Nova Scotians had the highest median age. And since then, Newfoundland & Labrador has had the highest median age of any Canadian province or territory.  

The chart below compares the median age in British Columbia in the years 1971 through 2015 to Newfoundland & Labrador. It has been a remarkable change there. For the years 1971 through 1993, it was the youngest province. Over the span of a dozen years, it went from the youngest to the oldest. And in 2015 the median age in Newfoundland & Labrador was 45--a full 4.5 years higher than the median age across Canada.

This dramatic shift in the age profile in Newfoundland & Labrador was a consequence of the collapse of the North Atlantic cod fishery; the moratorium instantly put 30,000 people out of work, and the population of the province fell by 10% between 1991 and 2001. [Source: Heritage Newfoundland & Labrador, "Rural Depopulation"](http://www.heritage.nf.ca/articles/society/depopulation.php)

```{r, echo=FALSE, message=FALSE}

# filter the long frame to include B.C. and Newfoundland and Labrador
regions <- c("British.Columbia", "Newfoundland.and.Labrador")
data_median_age_BC_prov <- 
  filter(data_medianage_long, region %in% regions)

# basic
plot_median_age_BC_prov <- 
  ggplot(data_median_age_BC_prov, aes(x=year, y=median_age, colour = region)) +
  geom_line(size = 1.5)

# fix the formatting
plot_median_age_BC_prov +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  ggtitle("Median age, British Columbia & Newfoundland & Labrador, 1971-2015") +
  ylab("median age") +
  scale_x_continuous(breaks = seq(1971, 2016, by = 5)) +
  scale_y_continuous(limits = c(20, 45), breaks = seq(20, 45, by = 5)) +
  theme(panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank())


```


-30-