---
title: "BC_demography_02_agedistribution"
author: "Martin Monkman"
date: "January 08, 2020"
output: html_document
---
<!--
Copyright 2020 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->


```{r setup, echo=FALSE, message_group=FALSE}

# tidyverse
library(tidyverse)

# utilities
library(glue)
library(here)
library(janitor)
#
# tables
library(xtable)
#
# plotting
library(gganimate)
library(zoo)
library(scales)
#

# PLOTTING ESSENTIALS

# create a chart theme & colour palette for use in ggplot2
# this theme is the basic background for a BC Stats chart
#
theme_BCStats <- 
  theme_bw() +
  theme(
    panel.border = element_rect(colour="white"),
    axis.line = element_line(colour="black"),
#    legend.position=c(1,0), 
#    legend.justification=c(1,0),
    legend.title = element_text(size=12),
    legend.text = element_text(size=11),
    axis.title = element_text(size=16),
    axis.text = element_text(size=16),
    plot.title = element_text(size=18)
  )
#

# colour palette for BC Stats charts (use with scale_colour_manual)
palette_BCStats <- c("#234275", "#E3A82B", "#26BDEF", "#11CC33", "#D3E2ED", "8A8A8A")

# grayscale for fill (use with scale_fill_manual)
palette_BCStats_fill <- c("#3F3F3F", "#ABABAB", "#DFDFDF", "#969696", "#838383", "8A8A8A")


#
txt_caption <- "Source: Statistics Canada, CANSIM Table 17-10-0005-01"
txt_caption2 <- "Source: Statistics Canada, CANSIM Tables 17-10-0005-01 & 17-10-0057-01"


```

## Plot the age distribution of British Columbia & Canada, 1971–present

### data


Source:
Statistics Canada, CANSIM 
[Population estimates on July 1st, by age and sex](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710000501)

Table: 17-10-0005-01 (formerly CANSIM 051-0001)

The source file contains population for Canada, the provinces and territories, by age (and age group), and by sex, for the years 1976-2020 (as of this writing).

Source file:  (https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710000501)


The CANSIM extract requires some carpentry to be useful for the purposes of this plotting. The tidied version becomes a data frame `data_age_groupdistrib_BC_2019`.

**Notes:** 

1. The table originally tabulated the last category as "90 years and over"; starting with 2001 this category was expanded to single years 90 through 99, and a "100 years and over" category was added. Depending on the year(s) of interest, the treatment of the extreme category will vary. 

2. The extreme age_group categories appear for all years, but are empty values. These lead to "parsing failures" when reading the source .csv file.


Read and clean data

```{r read_data, echo=FALSE, message_group=FALSE}

# read the data
df_1710000501 <- read_rds("../data_source/df_1710000501.rds")


```





Filter so that only single year categories remain (i.e. age 0, 1, 2, etc)

**Note:** the years up to and including 2000 include "90 years and over" as the oldest category; in 2001 the years 90-99 were included as single years, and the oldest category became "100 years and over". To make a data frame with single categories

* for those years, the 90+ category needs to be reframed as the single year "90"

* for the other years, the 90+ category needs to be dropped entirely

```{r}

# # getting the lay of the data
# ls.str(df_1710000501)
# 
# unique(df_1710000501$age_group)
# unique(df_1710000501$geo)
# unique(df_1710000501$ref_date)
# unique(df_1710000501$sex)
# 
# df_1710000501 %>% 
#   filter(sex == "Both sexes",
# #         age_group == "100 years and over", 
#          age_group == "99 years", 
#          geo == "Canada")



# age category values to drop
drop.age_group.vals_1 <- c("All ages", "0 to 4 years", "5 to 9 years",
                   "10 to 14 years", "15 to 19 years", 
                   "20 to 24 years", "25 to 29 years", 
                   "30 to 34 years", "35 to 39 years", 
                   "40 to 44 years", "45 to 49 years", 
                   "50 to 54 years", "55 to 59 years", 
                   "60 to 64 years", "65 to 69 years", 
                   "70 to 74 years", "75 to 79 years", 
                   "80 to 84 years", "85 to 89 years", 
                   "90 to 94 years", "95 to 99 years",
                   "0 to 14 years", "0 to 15 years", "0 to 16 years", "0 to 17 years",
                   "15 to 49 years", "15 to 64 years", "16 to 64 years", "17 to 64 years",
                   "18 years and over", "18 to 64 years", "18 to 24 years", "25 to 44 years",
                   "45 to 64 years", "65 years and over", 
                   "Median age", "Average age")

df_single <- df_1710000501 %>% 
  filter(!age_group %in% drop.age_group.vals_1)

df_single_1971 <- df_single %>% 
  filter(ref_date < 2001) %>% 
  mutate(age_group.num = gsub(" years", "", age_group)) %>% 
  mutate(age_group.num = gsub(" year", "", age_group.num)) %>% 
  mutate(age_group.num = gsub(" and over",  "", age_group.num)) %>% 
  mutate(age_group.num = as.integer(age_group.num))

df_single_2001 <- df_single %>% 
  filter(ref_date > 2000) %>% 
  filter(age_group != "90 years and over") %>% 
  mutate(age_group.num = gsub(" years", "", age_group)) %>% 
  mutate(age_group.num = gsub(" year", "", age_group.num)) %>% 
  mutate(age_group.num = gsub(" and over",  "", age_group.num)) %>% 
  mutate(age_group.num = as.integer(age_group.num))
  
data_age_groupdistrib_single <- rbind(df_single_1971, df_single_2001)  

#write_rds(data_age_groupdistrib_single, "../data_output/data_age_groupdistrib_single.rds")
#data_age_groupdistrib_single <- read_rds("../data_output/data_age_groupdistrib_single.rds")
```



```{r}

data_age_groupdistrib_single <- data_age_groupdistrib_single %>% 
  mutate(ref_date_year = as.integer(ref_date)) %>% 
  # add pct of year (by geo and sex)
  group_by(geo, ref_date, sex) %>% 
  mutate(year_pct = (value / sum(value) * 100)) %>% 
  ungroup()

# quick test
data_age_groupdistrib_single %>% 
  filter(geo == "British Columbia",
         ref_date == "2019",
         sex == "Both sexes") %>% 
  summarise(total_pop = sum(value),
            total_pct = sum(year_pct))

#write_rds(data_age_groupdistrib_single, "../data_output/data_age_groupdistrib_single.rds")
#data_age_groupdistrib_single <- read_rds("../data_output/data_age_groupdistrib_single.rds")


```




```{r}

# filter for British Columbia, both sexes, most recent year distribution

year_min <- min(data_age_groupdistrib_single$ref_date)
year_max <- max(data_age_groupdistrib_single$ref_date)

data_age_groupdistrib_BC_max <- data_age_groupdistrib_single %>%
  select(geo, ref_date, sex, age_group,  age_group.num, value, year_pct) %>%
  filter(geo == "British Columbia") %>%
  filter(ref_date == year_max) %>%
  filter(sex == "Both sexes")

#write_rds(data_age_groupdistrib_BC_max, "../data_source/data_age_groupdistrib_BC_max.rds")
#data_age_groupdistrib_BC_max <- read_rds("../data_source/data_age_groupdistrib_BC_max.rds")


```

### Plot: age_group distribution, British Columbia `r year_max`

This set of charts shows the distribution of the B.C. population, by one year age_group groups.

The first group shows the absolute number of people, and the second shows the percentage_group of the population in each group (useful for comparisons across time).


The first version of the plot is a basic version. The second version adds an assortment of formatting:

* axis length and grid specifications
* axis titles
* chart title, subtitle, and source

Note that the X axis scale is such that it follows the five year increments of the Census of Canada.



```{r}

# basic
plot_age_groupdist_BC_max <- ggplot(data_age_groupdistrib_BC_max, aes(x=age_group.num, y=value)) +
  geom_bar(stat="identity", fill="#234275")

plot_age_groupdist_BC_max


# Add formatting

plot_age_groupdist_BC_max <- plot_age_groupdist_BC_max +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  labs(title = "British Columbia: age distribution (total)",
       subtitle = year_max,
       y = "population", x = "age",
       caption = txt_caption) +
  scale_x_continuous(breaks = seq(0, 100, 10)) +
#  scale_x_discrete(breaks = c("10", "20", "30", "40", "50", 
#                              "60", "70", "80", "90", "100+")) +
  scale_y_continuous(limits = c(0, 90000), breaks = seq(0, 90000, by = 10000),
                     labels = comma)

plot_age_groupdist_BC_max

## save the plot as a png file
plot_file_name <- glue('../plot_images/plot_age_groupdist_BC_{year_max}.png')
ggsave(plot_file_name, width=9, height=6)
#

```


Repeat, but with percentage_group

```{r}

# basic
plot_age_groupdist_BC_max_pct <- ggplot(data_age_groupdistrib_BC_max, aes(x=age_group.num, y=year_pct)) +
  geom_bar(stat="identity", fill="#234275")

plot_age_groupdist_BC_max_pct


# Add formatting

plot_age_groupdist_BC_max_pct <- plot_age_groupdist_BC_max_pct +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  labs(title = "British Columbia: age distribution (percentage)",
       subtitle = year_max,
       y = "population", x = "age",
       caption = txt_caption) +
  scale_x_continuous(breaks = seq(0, 100, 10)) +
  scale_y_continuous(limits = c(0, 2.25), breaks = seq(0, 2.0, by = 0.5)) 

plot_age_groupdist_BC_max_pct


## save the plot as a png file
plot_file_name <- glue('../plot_images/plot_age_groupdist_BC_{year_max}_pct.png')
ggsave(plot_file_name, width=9, height=6)

```


```{r}

# highlight the generations

# baby boom (in Canada): 1946 - 1965
# gen X: 1966 - 1980
# millennials: 1981 - 1995
# gen X: 1996 - 2001

# https://www.pewresearch.org/fact-tank/2019/01/17/where-millennials-end-and-generation-z-begins/

boomer_max <- as.integer(year_max) - 1946
boomer_min <- as.integer(year_max) - 1965
genx_max <- as.integer(year_max) - 1965
genx_min <- as.integer(year_max) - 1980
millen_max <- as.integer(year_max) - 1980
millen_min <- as.integer(year_max) - 1995
genz_max <- as.integer(year_max) - 1995
genz_min <- as.integer(year_max) - 2010


plot_age_groupdist_BC_max_shade <- plot_age_groupdist_BC_max +
  # add shaded rectangle for the projection range
  # boomer
  annotate("rect", xmin=boomer_min, xmax=boomer_max, ymin=0, ymax=89750, alpha=0.1, fill="green") +
  annotate("text", x=(boomer_min + 7), y=85000,  label="Boomers") +
  # gen X
  annotate("rect", xmin=genx_min, xmax=genx_max, ymin=0, ymax=89750, alpha=0.1, fill="blue") +
  annotate("text", x=(genx_min + 5), y=85000,  label="GenX") +
  # avocado toast
  annotate("rect", xmin=millen_min, xmax=millen_max, ymin=0, ymax=89750, alpha=0.1, fill="red") +
  annotate("text", x=(millen_min + 7), y=85000,  label="Millennials") +
  # post-avocado toast
  annotate("rect", xmin=genz_min, xmax=genz_max, ymin=0, ymax=89750, alpha=0.1, fill="orange") +
  annotate("text", x=(genz_min + 5), y=85000,  label="GenZ") 
  

plot_age_groupdist_BC_max_shade

# save the plot as a png file
# save the plot as a png file
plot_file_name <- glue('../plot_images/plot_age_groupdist_BC_{year_max}_shade.png')
ggsave(plot_file_name, width=9, height=6)



```


The total population of British Columbia has increased from 2.2 million in 1971 to 5.1 million in 2019. The population surpassed 3 million in 1986, and 4 million in 1999.


#### Density Plot

**NOTE**

Not really a density plot at all! A density plot is essentially an extension of a histogram calculation; here the histogram is pre-calculated, so that the solution is a loess smoothing line.  Some eye-balling suggests that a span = 0.3 is a pretty good approximation (see below).


Having two smoothed lines representing the top of the bars is an effective way to compare the distribution of the population at two or three points in time.



Plot with curves

```{r}

plot_age_groupdist_BC_max +
 geom_smooth(span = 0.3, se = FALSE, colour = "#E3A82B", size = 1.5)  

# save the plot as a png file
ggsave("../plot_images/plot_age_groupdist_BC_max_smooth.png", width=9, height=6)
  

```


```{r}
# regular plot -- most recent year

plot_smooth_years <- data_age_groupdistrib_single %>% 
  filter(geo == "British Columbia" & sex == "Both sexes") %>% 
  filter(ref_date_year == year_min | ref_date_year == year_max) %>% 
  ggplot(aes(x=age_group.num, y=value, colour = ref_date)) +
  geom_smooth(span = 0.3, se = FALSE)

plot_smooth_years

plot_smooth_years <- plot_smooth_years +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  labs(title = "British Columbia: age distribution (total)",
       subtitle = glue('{year_min} & {year_max}'),
       y = "population", x = "age",
       caption = txt_caption) +
  scale_x_continuous(breaks = seq(0, 100, 10)) +
  scale_y_continuous(limits = c(0, 90000), breaks = seq(0, 90000, by = 10000),
                     labels = comma) +
#  theme(
#    legend.position = c(1, 1),
#    legend.justification = c("right", "top"),
#    legend.title = element_blank()
#  ) +
  theme(legend.position = "none") +
  annotate("text", x=25, y=30000, label = year_min, colour =  "#234275", fontface = 2) +
  annotate("text", x=64, y=60000, label = year_max, colour =  "#234275", fontface = 2)
  
  

plot_smooth_years

# save the plot as a png file
ggsave("../plot_images/plot_smooth_years.png", width=9, height=6)
  
  


```


## animate


```{r eval=FALSE}

p <- data_age_groupdistrib_single %>% 
  filter(geo == "British Columbia" & sex == "Both sexes") %>% 
  #
  ggplot(aes(x=age_group.num, y=value)) +
  geom_bar(stat="identity", fill="#234275") +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  labs(title = "British Columbia: age distribution (total)",
#       subtitle = "2019",
       y = "population", x = "age",
       caption = txt_caption) +
  scale_x_continuous(breaks = seq(0, 100, 10)) +
#  scale_x_discrete(breaks = c("10", "20", "30", "40", "50", 
#                              "60", "70", "80", "90", "100+")) +
  scale_y_continuous(limits = c(0, 90000), breaks = seq(0, 90000, by = 10000),
                     labels = comma) +
  # gganimate
#  transition_time(ref_date_year) +
  transition_time(ref_date_year) +
  labs(subtitle = "Year: {frame_time}")


#p
#
#animate(p, duration = 30)
#
#anim_save("BC_pop_dist_anim.gif")

```


```{r}
# combined plot: percentage_group

plot_smooth_pct <- 
  data_age_groupdistrib_single %>% 
  filter(geo == "British Columbia" & sex == "Both sexes") %>% 
  filter(ref_date_year == year_min | ref_date_year == year_max) %>% 
  ggplot(aes(x=age_group.num, y=year_pct, colour = ref_date)) +
  geom_smooth(span = 0.3, se = FALSE) + 
  annotate("text", x=20, y=2.0, label = year_min, colour =  "#234275", fontface = 2) +
  annotate("text", x=66, y=1.5, label = year_max, colour =  "#234275", fontface = 2)
  

plot_smooth_pct <- plot_smooth_pct +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  labs(title = "British Columbia: age distribution (percent)",
       subtitle = glue('{year_min} & {year_max}'),
       y = "percent of population", x = "age",
       caption ="Source: Statistics Canada, CANSIM Table 051-0001") +
  scale_x_continuous(breaks = seq(0, 100, 10)) +
  scale_y_continuous(limits = c(0, 2.25), breaks = seq(0, 2.0, by = 0.5)) +
#  theme(
#    legend.position = c(1, 1),
#    legend.justification = c("right", "top"),
#    legend.title = element_blank()
#  ) +
  theme(legend.position = "none")
  
  
  
plot_smooth_pct

## save the plot as a png file
#png("plot_smooth_pct.png", width=1024, height=768, res=120)
#plot(plot_smooth_pct)
#dev.off()
  

```

## Projections


Population Projections for Canada (2018 to 2068), Provinces and Territories (2018 to 2043)

https://www150.statcan.gc.ca/n1/pub/91-520-x/91-520-x2019001-eng.htm

Data: 

Projected population, by projection scenario, age and sex, as of July 1 (x 1,000)1
Frequency: Annual

Table: 17-10-0057-01 (formerly CANSIM 052-0005)

Geography: Canada, Province or territory

https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710005701



**NOTE**

This plot uses the "M3" projection:

> The medium-growth (M3) scenario contains the following assumptions at the Canada level: the total fertility rate reaches 1.59 children per woman in 2042/2043 and remains constant thereafter; life expectancy at birth reaches 87.0 years for males and 89.0 years for females in 2067/2068; interprovincial migration is based on the trends observed between 2003/2004 and 2008/2009; the immigration rate reaches 0.83% in 2042/2043 and remains constant thereafter; the annual number of non-permanent residents reaches 1,397,060 in 2043 and remains constant thereafter; the net emigration rate reaches 0.15% in 2042/2043 and remains constant thereafter.




```{r read_data_projection, echo=FALSE, message_group=FALSE}

# read the data
data_BC_projection <- read_rds("../data_output/data_BC_projection.rds")


```


```{r}
df_single_proj <- data_BC_projection %>% 
  filter(!age_group %in% drop.age_group.vals_1 &
           sex == "Both sexes" &
           ref_date > year_max) %>% 
  # drop rows for the years that StatCan only projects Canadian population
  drop_na(value) %>% 
  mutate(value_th = value * 1000)
  

# create new variable with just the number component of the age group string
df_single_proj <- df_single_proj %>% 
  mutate(age_group.num = 
           case_when(
             age_group == "Under 1 year" ~ 0,
             TRUE ~ parse_number(age_group))
  )

```



```{r}

year_max_proj <- max(df_single_proj$ref_date)

df_single_proj_max <- df_single_proj %>% 
  group_by(ref_date) %>% 
  mutate(age_grp_pct = value / sum(value) * 100) %>% 
  filter(ref_date == year_max_proj)

df_single_proj_max

```


```{r}

plot_smooth_years_proj <-
plot_smooth_years + 
  geom_smooth(data = df_single_proj_max, 
              aes(x = age_group.num, y = value_th),
              span = 0.3, se = FALSE) +
  annotate("text", x=90, y=50000, label = year_max_proj, colour =  "#234275", fontface = 2) +
  labs(subtitle = glue('{year_min}, {year_max} & {year_max_proj} (M3 projection)'),
       caption = txt_caption2)  +
  theme(legend.position = "none")
plot_smooth_years_proj

# save the plot as a png file
ggsave("../plot_images/plot_smooth_years_proj.png", width=9, height=6)
  
  
plot_smooth_pct_proj <-
plot_smooth_pct +
  geom_smooth(data = df_single_proj_max, 
              aes(x = age_group.num, y = age_grp_pct),
              span = 0.3, se = FALSE) +
  annotate("text", x=86, y=1.0, label = year_max_proj, colour =  "#234275", fontface = 2) +
  labs(subtitle = glue('{year_min}, {year_max} & {year_max_proj} (M3 projection)'),
       caption = txt_caption2) +
  theme(legend.position = "none")

plot_smooth_pct_proj

# save the plot as a png file
ggsave("../plot_images/plot_smooth_pct_proj.png", width=9, height=6)

  

```


## 65+ and 85+

```{r}

#data_age_groupdistrib_single <- read_rds("../data_output/data_age_groupdistrib_single.rds")
data_age_groupdistrib_single

df_65plus_BC <- 
data_age_groupdistrib_single %>% 
  filter(geo == "British Columbia",
         sex == "Both sexes",
         age_group.num >= 65)

df_65plus_BC <- df_65plus_BC %>% 
  select(ref_date_year, geo, sex, age_group, age_group.num, value)

df_65plus_BC

```


```{r}
df_single_proj

df_single_proj_65 <- df_single_proj %>% 
  mutate(ref_date_year = as.integer(ref_date)) %>% 
  filter(age_group.num >= 65) %>% 
  select(ref_date_year, geo, sex, age_group, age_group.num, value = value_th) 

df_single_proj_65

df_65plus_BC_proj <- rbind(df_65plus_BC, df_single_proj_65)

tail(df_65plus_BC_proj)
```



```{r}
df_65plus_summary <-
df_65plus_BC_proj %>% 
  drop_na(value) %>% 
  group_by(geo, ref_date_year, sex) %>% 
  summarise(total_pop_65 = sum(value),
            total_pop_80 = sum(value[age_group.num >= 80]))

df_65plus_summary
```

#### chart

```{r}

ch_65plus <- 
ggplot(df_65plus_summary) +
  geom_line(aes(x = ref_date_year, y = total_pop_65), colour = "#234275", size = 1.5) +
  geom_line(aes(x = ref_date_year, y = total_pop_80), colour = "#E3A82B", size = 1.5) 
  
ch_65plus

```



```{r}
ch_65plus +
  theme_BCStats +
#  scale_colour_manual(values=palette_BCStats) +
  labs(title = "British Columbia: population, ages 65+ & 80+",
       subtitle = glue('{year_min}–{year_max_proj}'),
       y = "population", x = "",
       caption = txt_caption2) +
#  scale_x_continuous(breaks = seq(0, 100, 10)) +
  scale_y_continuous(limits = c(0, 1750000), breaks = seq(0, 1750000, by = 250000),
                     labels = comma) +
#  theme(
#    legend.position = c(1, 1),
#    legend.justification = c("right", "top"),
#    legend.title = element_blank()
#  ) +
  theme(legend.position = "none") +
  annotate("text", x=1990, y=500000, label = "65+", colour =  "#234275", fontface = 2) +
  annotate("text", x=2015, y=300000, label = "80+", colour =  "#E3A82B", fontface = 2) +
#  
  # projection box
  annotate("rect", xmin=as.numeric(year_max), xmax=as.numeric(year_max_proj), ymin=0, ymax=1750000, alpha=0.1, fill="blue") +
  annotate("text", x=(as.numeric(year_max) + 5), y=1650000,  label="projection", colour = "blue") 
  

# save the plot as a png file
ggsave("../plot_images/plot_65plus.png", width=9, height=6)


```


## Summary statistics

Questions:

- what percentage_group of British Columbians are 65 and older?
- how many school-age_group children are there?
- when was the peak in school age_group children?

-30-
