---
title: "BC_demography_05_lifeexpectancy"
author: "Martin Monkman"
date: "February 26, 2016"
output: html_document
---
<!--
Copyright 2020 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->



```{r, echo=FALSE, message=FALSE}

# THE PRELIMINARIES

# load the necessary packages
# data manipulation
library(tidyr)
library(dplyr)
# graphics
library(ggplot2)
library(scales)
library(RColorBrewer)
# tables
library(xtable)


# PLOTTING ESSENTIALS

# create a chart theme & colour palette for use in ggplot2
# this theme is the basic background for a BC Stats chart
#
theme_BCStats <- 
  theme_bw() +
  theme(
    panel.border = element_rect(colour="white"),
    axis.line = element_line(colour="black"),
    legend.position=c(1,0), 
    legend.justification=c(1,0),
    legend.title = element_text(size=12),
    legend.text = element_text(size=11),
    axis.title = element_text(size=12),
    axis.text = element_text(size=12)
  )
#

# colour palette for BC Stats charts (use with scale_colour_manual)
palette_BCStats <- c("#234275", "#E3A82B", "#26BDEF", "#11CC33", "#D3E2ED", "8A8A8A")

# grayscale for fill (use with scale_fill_manual)
palette_BCStats_fill <- c("#3F3F3F", "#ABABAB", "#DFDFDF", "#969696", "#838383", "8A8A8A")

# NOTE: the RColorBrewer palette "Greys" can be used in order to provide more shades

```



## Plot of the change in life expectancy in British Columbia, 1971-2015

### data

Summary table: [Life expectancy at birth, by sex, by province](http://www.statcan.gc.ca/tables-tableaux/sum-som/l01/cst01/health26-eng.htm)

Life tables: [Life Tables, Canada, Provinces and Territories (84-537-X)](http://www5.statcan.gc.ca/olc-cel/olc.action?objId=84-537-X&objType=2&lang=en&limit=0)

Data source: [Life expectancy, at birth and at age 65, by sex, Canada, provinces and territories; CANSIM Table 102-0512](http://www5.statcan.gc.ca/cansim/a26?lang=eng&retrLang=eng&id=1020512&&pattern=&stByVal=1&p1=1&p2=37&tabMode=dataTable&csid=)


The CANSIM Table 102-0512 was used to download the file:  (data_source/cansim-1020512-eng-5562131734164497858.csv)


The CANSIM extract was cleaned up to produce a wide table that contained only the Geography (as the first variable), Sex, and Life Expectancy category (At birth & At age 65). 

A CSV of this sheet was saved as BC_lifeexpectancy.csv, which is read into the R workspace.

The resulting table is then melted into long format.

#### Defining "life expectancy"

From [footnote 3 in the CANSIM table 102-0512](http://www5.statcan.gc.ca/cansim/a26?lang=eng&retrLang=eng&id=1020512&&pattern=&stByVal=1&p1=1&p2=37&tabMode=dataTable&csid=):

> Life expectancy is the number of years a person would be expected to live, 
> starting at birth (for life expectancy at birth) or at age 65 (for life expectancy at age 65) 
> if the age- and sex-specific mortality rates for a given observation period 
> (such as a calendar year) were held constant over his/her life span.


```{r, echo=FALSE, message=FALSE}



# read the data
setwd("C:/@data/GitHub/BC_demography")
BC_lifeexpectancy <- read.csv("./data_source/BC_lifeexpectancy.csv", header = TRUE)
#data_medianage <- read.csv("data_medianage.csv", header = TRUE)

# turn the wide version of data_medianage into long format
BC_lifeexpectancy_long <- BC_lifeexpectancy %>%
  gather(Year, Life_expectancy, -Sex, -Geography, -Life.expectancy)

head(BC_lifeexpectancy_long)

# filter the long frame to include only B.C. 
#data_median_age_BC <- filter(data_medianage_long, region == "British.Columbia")

```

Note to self: the source data could/should just be the entire national field, and then filtered to include only the B.C. series.



### B.C. life expectancy plot -- from At birth

The first version plots the change in life expectancy at birth.






```{r, echo=FALSE, message=FALSE}


#Note that the X axis scale is such that it follows the five year increments of the Census of Canada.
#OK maybe in this case it doesn't

# filter to include just "At birth" and "Both sexes"

BC_lifeexpectancy_Atbirth <- BC_lifeexpectancy_long %>%
  filter(Life.expectancy == "At birth", Sex == "Both sexes")

head(BC_lifeexpectancy_Atbirth)

# basic
plot_lifeexpectancy_BC <- 
  ggplot(BC_lifeexpectancy_Atbirth, aes(x=Year, y=Life_expectancy, group=Geography)) +
  geom_line(size = 1.5)

plot_lifeexpectancy_BC


# fix the formatting


plot_lifeexpectancy_BC +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  ggtitle("British Columbia, Life expectancy at birth, 1992-2009") +
  ylab("life expectancy") +
  #scale_x_discrete(breaks = seq(1991, 2012, by = 5)) +
  scale_y_continuous(limits = c(75, 85), breaks = seq(75, 85, by = 1)) +
  # remove the legend (not much point when there's only one line!)
  theme(legend.position="none")




```


Fix the X axis values



Add the following:

* Life expectancy at age 65

* compare men and women (closing the gap)

Is there data going further back?

-30-