---
title: "BC_demography_02_agedistribution"
author: "Martin Monkman"
date: "January 08, 2020"
output: html_document
---
<!--
Copyright 2020 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->


## Plot the age distribution of British Columbia & Canada, 1971–present

### data


Source:
Statistics Canada, CANSIM 
[Population estimates on July 1st, by age and sex](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710000501)

Table: 17-10-0005-01 (formerly CANSIM 051-0001)

The source file contains population for Canada, the provinces and territories, by age (and age group), and by sex, for the years 1976-2020 (as of this writing).

Source file:  (https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710000501)


The CANSIM extract requires some carpentry to be useful for the purposes of this plotting. The tidied version becomes a data frame `data_age_groupdistrib_BC_2019`.

**Notes:** 

1. The table originally tabulated the last category as "90 years and over"; starting with 2001 this category was expanded to single years 90 through 99, and a "100 years and over" category was added. Depending on the year(s) of interest, the treatment of the extreme category will vary. 

2. The extreme age_group categories appear for all years, but are empty values. These lead to "parsing failures" when reading the source .csv file.


```{r setup, echo=FALSE, message_group=FALSE}

# tidyverse
library(tidyverse)

# utilities
library(here)
library(janitor)
#
# tables
library(xtable)
#
# plotting
library(gganimate)
library(zoo)
library(scales)
#
library(cansim)

# PLOTTING ESSENTIALS

# create a chart theme & colour palette for use in ggplot2
# this theme is the basic background for a BC Stats chart
#
theme_BCStats <- 
  theme_bw() +
  theme(
    panel.border = element_rect(colour="white"),
    axis.line = element_line(colour="black"),
#    legend.position=c(1,0), 
#    legend.justification=c(1,0),
    legend.title = element_text(size=12),
    legend.text = element_text(size=11),
    axis.title = element_text(size=16),
    axis.text = element_text(size=16),
    plot.title = element_text(size=18)
  )
#

# colour palette for BC Stats charts (use with scale_colour_manual)
palette_BCStats <- c("#234275", "#E3A82B", "#26BDEF", "#11CC33", "#D3E2ED", "8A8A8A")

# grayscale for fill (use with scale_fill_manual)
palette_BCStats_fill <- c("#3F3F3F", "#ABABAB", "#DFDFDF", "#969696", "#838383", "8A8A8A")


#
txt_caption <- "Source: Statistics Canada, CANSIM Table 17-10-0005-01"


```

Read and clean data

```{r read_data, echo=FALSE, message_group=FALSE}

# read the data
df_1710000501 <- get_cansim("17-10-0005-01")

df_1710000501 <- clean_names(df_1710000501)




```





Filter so that only single year categories remain (i.e. age 0, 1, 2, etc)

**Note:** the years up to and including 2000 include "90 years and over" as the oldest category; in 2001 the years 90-99 were included as single years, and the oldest category became "100 years and over". To make a data frame with single categories

* for those years, the 90+ category needs to be reframed as the single year "90"

* for the other years, the 90+ category needs to be dropped entirely

```{r}

# # getting the lay of the data
# ls.str(df_1710000501)
# 
# unique(df_1710000501$age_group)
# unique(df_1710000501$geo)
# unique(df_1710000501$ref_date)
# unique(df_1710000501$sex)
# 
# df_1710000501 %>% 
#   filter(sex == "Both sexes",
# #         age_group == "100 years and over", 
#          age_group == "99 years", 
#          geo == "Canada")



# age category values to drop
drop.age_group.vals_1 <- c("All ages", "0 to 4 years", "5 to 9 years",
                   "10 to 14 years", "15 to 19 years", 
                   "20 to 24 years", "25 to 29 years", 
                   "30 to 34 years", "35 to 39 years", 
                   "40 to 44 years", "45 to 49 years", 
                   "50 to 54 years", "55 to 59 years", 
                   "60 to 64 years", "65 to 69 years", 
                   "70 to 74 years", "75 to 79 years", 
                   "80 to 84 years", "85 to 89 years", 
                   "90 to 94 years", "95 to 99 years",
                   "0 to 14 years", "0 to 15 years", "0 to 16 years", "0 to 17 years",
                   "15 to 49 years", "15 to 64 years", "16 to 64 years", "17 to 64 years",
                   "18 years and over", "18 to 64 years", "18 to 24 years", "25 to 44 years",
                   "45 to 64 years", "65 years and over", 
                   "Median age")

df_single <- df_1710000501 %>% 
  filter(!age_group %in% drop.age_group.vals_1)

df_single_1971 <- df_single %>% 
  filter(ref_date < 2001) %>% 
  mutate(age_group.num = gsub(" years", "", age_group)) %>% 
  mutate(age_group.num = gsub(" year", "", age_group.num)) %>% 
  mutate(age_group.num = gsub(" and over",  "", age_group.num)) %>% 
  mutate(age_group.num = as.integer(age_group.num))

df_single_2001 <- df_single %>% 
  filter(ref_date > 2000) %>% 
  filter(age_group != "90 years and over") %>% 
  mutate(age_group.num = gsub(" years", "", age_group)) %>% 
  mutate(age_group.num = gsub(" year", "", age_group.num)) %>% 
  mutate(age_group.num = gsub(" and over",  "", age_group.num)) %>% 
  mutate(age_group.num = as.integer(age_group.num))
  
data_age_groupdistrib_single <- rbind(df_single_1971, df_single_2001)  


```



```{r}

data_age_groupdistrib_single <- data_age_groupdistrib_single %>% 
  mutate(ref_date_year = as.integer(ref_date)) %>% 
  # add pct of year (by geo and sex)
  group_by(geo, ref_date, sex) %>% 
  mutate(year_pct = (value / sum(value) * 100)) %>% 
  ungroup()

# quick test
data_age_groupdistrib_single %>% 
  filter(geo == "British Columbia",
         ref_date == "2019",
         sex == "Both sexes") %>% 
  summarise(total_pop = sum(value),
            total_pct = sum(year_pct))

```




```{r}

# filter for British Columbia, both sexes, most recent year distribution

year_min <- min(data_age_groupdistrib_single$ref_date)
year_max <- max(data_age_groupdistrib_single$ref_date)

data_age_groupdistrib_BC_max <- data_age_groupdistrib_single %>%
  select(geo, ref_date, sex, age_group,  age_group.num, value, year_pct) %>%
  filter(geo == "British Columbia") %>%
  filter(ref_date == year_max) %>%
  filter(sex == "Both sexes")

```

### Plot: age_group distribution, British Columbia `r year_max`

This set of charts shows the distribution of the B.C. population, by one year age_group groups.

The first group shows the absolute number of people, and the second shows the percentage_group of the population in each group (useful for comparisons across time).


The first version of the plot is a basic version. The second version adds an assortment of formatting:

* axis length and grid specifications
* axis titles
* chart title, subtitle, and source

Note that the X axis scale is such that it follows the five year increments of the Census of Canada.



```{r}

# basic
plot_age_groupdist_BC_max <- ggplot(data_age_groupdistrib_BC_max, aes(x=age_group.num, y=value)) +
  geom_bar(stat="identity", fill="#234275")

plot_age_groupdist_BC_max


# Add formatting

plot_age_groupdist_BC_max <- plot_age_groupdist_BC_max +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  labs(title = "British Columbia: age distribution (total)",
       subtitle = year_max,
       y = "population", x = "age",
       caption = txt_caption) +
  scale_x_continuous(breaks = seq(0, 100, 10)) +
#  scale_x_discrete(breaks = c("10", "20", "30", "40", "50", 
#                              "60", "70", "80", "90", "100+")) +
  scale_y_continuous(limits = c(0, 90000), breaks = seq(0, 90000, by = 10000),
                     labels = comma)

plot_age_groupdist_BC_max

# save the plot as a png file
plot_file_name <- glue('plot_age_groupdist_BC_{year_max}.png')
ggsave(plot_file_name, width=9, height=6)


```


Repeat, but with percentage_group

```{r}

# basic
plot_age_groupdist_BC_max_pct <- ggplot(data_age_groupdistrib_BC_max, aes(x=age_group.num, y=year_pct)) +
  geom_bar(stat="identity", fill="#234275")

plot_age_groupdist_BC_max_pct


# Add formatting

plot_age_groupdist_BC_max_pct <- plot_age_groupdist_BC_max_pct +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  labs(title = "British Columbia: age distribution (percentage)",
       subtitle = year_max,
       y = "population", x = "age",
       caption = txt_caption) +
  scale_x_continuous(breaks = seq(0, 100, 10)) +
  scale_y_continuous(limits = c(0, 2.25), breaks = seq(0, 2.0, by = 0.5)) 

plot_age_groupdist_BC_max_pct


# save the plot as a png file
plot_file_name <- glue('plot_age_groupdist_BC_{year_max}_pct.png')
ggsave(plot_file_name, width=9, height=6)

```


```{r}

# highlight the generations

# baby boom (in Canada): 1946 - 1965
# gen X: 1966 - 1980
# millenials: 1981 - 1995


boomer_max <- as.integer(year_max) - 1946
boomer_min <- as.integer(year_max) - 1965
genx_max <- as.integer(year_max) - 1965
genx_min <- as.integer(year_max) - 1980
millen_max <- as.integer(year_max) - 1980
millen_min <- as.integer(year_max) - 1995


plot_age_groupdist_BC_max_shade <- plot_age_groupdist_BC_max +
  # add shaded rectangle for the projection range
  # boomer
  annotate("rect", xmin=boomer_min, xmax=boomer_max, ymin=0, ymax=89750, alpha=0.1, fill="green") +
  annotate("text", x=(boomer_min + 6), y=85000,  label="Boomers") +
  # gen X
  annotate("rect", xmin=genx_min, xmax=genx_max, ymin=0, ymax=89750, alpha=0.1, fill="blue") +
  annotate("text", x=(genx_min + 5), y=85000,  label="GenX") +
  # avocado toast
  annotate("rect", xmin=millen_min, xmax=millen_max, ymin=0, ymax=89750, alpha=0.1, fill="red") +
  annotate("text", x=(millen_min + 6), y=85000,  label="Millenials") 
  

plot_age_groupdist_BC_max_shade

# save the plot as a png file
# save the plot as a png file
plot_file_name <- glue('plot_age_groupdist_BC_{year_max}_shade.png')
ggsave(plot_file_name, width=9, height=6)



```


The total population of British Columbia has increased from 2.2 million in 1971 to 5.1 million in 2019. The population surpassed 3 million in 1986, and 4 million in 1999.


#### Density Plot

**NOTE**

Not really a density plot at all! A density plot is essentially an extension of a histogram calculation; here the histogram is pre-calculated, so that the solution is a loess smoothing line.  Some eye-balling suggests that a span = 0.3 is a pretty good approximation (see below).


Having two smoothed lines representing the top of the bars is an effective way to compare the distribution of the population at two or three points in time.



Plot with curves

```{r}
# regular plot -- most recent year

plot_smooth_years <- data_age_groupdistrib_single %>% 
  filter(geo == "British Columbia" & sex == "Both sexes") %>% 
  filter(ref_date_year == year_min | ref_date_year == year_max) %>% 
  ggplot(aes(x=age_group.num, y=value, colour = ref_date)) +
  geom_smooth(span = 0.3, se = FALSE)

plot_smooth_years

plot_smooth_years <- plot_smooth_years +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  labs(title = "British Columbia: age distribution (total)",
       subtitle = glue('{year_min} & {year_max}'),
       y = "population", x = "age",
       caption = txt_caption) +
  scale_x_continuous(breaks = seq(0, 100, 10)) +
  scale_y_continuous(limits = c(0, 90000), breaks = seq(0, 90000, by = 10000),
                     labels = comma) +
  theme(
    legend.position = c(1, 1),
    legend.justification = c("right", "top"),
    legend.title = element_blank()
  )

plot_smooth_years

# save the plot as a png file
png("plot_smooth_years.png", width=1024, height=768, res=120)
plot(plot_smooth_years)
dev.off()
  


```


## animate


```{r eval=FALSE}

p <- data_age_groupdistrib_single %>% 
  filter(geo == "British Columbia" & sex == "Both sexes") %>% 
  #
  ggplot(aes(x=age_group.num, y=value)) +
  geom_bar(stat="identity", fill="#234275") +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  labs(title = "British Columbia: age distribution (total)",
#       subtitle = "2019",
       y = "population", x = "age",
       caption = txt_caption) +
  scale_x_continuous(breaks = seq(0, 100, 10)) +
#  scale_x_discrete(breaks = c("10", "20", "30", "40", "50", 
#                              "60", "70", "80", "90", "100+")) +
  scale_y_continuous(limits = c(0, 90000), breaks = seq(0, 90000, by = 10000),
                     labels = comma) +
  # gganimate
#  transition_time(ref_date_year) +
  transition_time(ref_date_year) +
  labs(subtitle = "Year: {frame_time}")


p

animate(p, duration = 30)

anim_save("BC_pop_dist_anim.gif")

```


## _RESUME HERE_



```{r}


# combined plot: percentage_group
plot_smooth_pct <- ggplot(data_age_groupdistrib_BC_2019, aes(x=age_group.num, y=year_pct)) +
  geom_smooth(span = 0.3, se = FALSE, colour = "#234275") +
  geom_smooth(data = data_age_groupdistrib_BC_1971, span = 0.3, se = FALSE, colour = "#E3A82B") +
  annotate("text", x=20, y=2.0, label = "1971", colour =  "#234275", fontface = 2) +
  annotate("text", x=66, y=1.5, label = "2016", colour =  "#234275", fontface = 2)
  

plot_smooth_pct <- plot_smooth_pct +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  labs(title = "British Columbia: age_group distribution (percent)",
       subtitle = "1971 & 2016",
       y = "percent of population", x = "age_group",
       caption ="Source: Statistics Canada, CANSIM Table 051-0001") +
  scale_x_continuous(breaks = seq(0, 100, 10)) +
  scale_y_continuous(limits = c(0, 2.25), breaks = seq(0, 2.0, by = 0.5)) 
  
plot_smooth_pct

# save the plot as a png file
png("plot_smooth_pct.png", width=1024, height=768, res=120)
plot(plot_smooth_pct)
dev.off()
  

```


## Add 2041 projection to smoothed curve plots

#### read the data

```{r}

data_BC_projection <- read_csv("./data_source/Project_Inputs_TBS_Proj1607_5_single-all.csv")

ls.str(data_BC_projection)

unique(data_BC_projection$age_group)
unique(data_BC_projection$sex)
unique(data_BC_projection$Year)

data_BC_projection %>%
  group_by(sex) %>%
  count()

# delete rows where sex==NA

data_BC_projection <- data_BC_projection %>%
  na.omit()

data_BC_projection %>%
  group_by(sex) %>%
  count()

unique(data_BC_projection$age_group)
unique(data_BC_projection$sex)
unique(data_BC_projection$Year)


# clean up age_group category labels
data_BC_projection$age_group.num <- as.integer(data_BC_projection$age_group) 
data_BC_projection$age_group.num[data_BC_projection$age_group == "100+"] <- 100 

data_BC_projection %>%
  group_by(sex) %>%
  count()

unique(data_BC_projection$age_group.num)

ls.str(data_BC_projection)

data_age_groupdistrib_BC_2041 <- data_BC_projection %>%
#  select(geo, ref_date, sex, age_group,  value) %>%
  filter(Year == "2041") %>%
  group_by(age_group.num) %>%
  summarise(value = sum(Population))

ls.str(data_age_groupdistrib_BC_2041)


# calculate percentage_group of total population for each age_group group
data_age_groupdistrib_BC_2041 <- data_age_groupdistrib_BC_2041 %>%
  mutate(year_pct = (value / sum(value) * 100)) 

ls.str(data_age_groupdistrib_BC_2041)

```


#### Plot


```{r}

plot_age_groupdist_BC_2041 <- ggplot(data_age_groupdistrib_BC_2041, aes(x=age_group.num, y=value)) +
  geom_bar(stat="identity", fill="#234275")

plot_age_groupdist_BC_2041
  

ggplot(data_age_groupdistrib_BC_2041, aes(x=age_group.num, y=value)) +
  geom_bar(stat="identity", fill="#234275") +
  geom_smooth(span = 0.25, se = FALSE)
  
  


# combined plot: absolute
plot_smooth_years2 <- ggplot(data_age_groupdistrib_BC_2019, aes(x=age_group.num, y=value)) +
  geom_smooth(span = 0.25, se = FALSE, colour = "#234275") +
  geom_smooth(data = data_age_groupdistrib_BC_1971, span = 0.25, se = FALSE, colour = "#E3A82B") +
  geom_smooth(data = data_age_groupdistrib_BC_2041, span = 0.25, se = FALSE, colour = "#26BDEF") +
  annotate("text", x=19, y=35000, label = "1971", colour =  "#234275", fontface = 2) +
  annotate("text", x=68, y=35000, label = "2016", colour =  "#234275", fontface = 2) +
  annotate("text", x=93, y=35000, label = "2041", colour =  "#234275", fontface = 2)

plot_smooth_years2

plot_smooth_years2 <- plot_smooth_years2 +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  labs(title = "British Columbia: age_group distribution (total)",
       subtitle = "1971, 2016 & 2041",
       y = "population", x = "age_group",
       caption ="Source: Statistics Canada, CANSIM Table 051-0001; BC Stats") +
  scale_x_continuous(breaks = seq(0, 100, 10)) +
  scale_y_continuous(limits = c(0, 90000), breaks = seq(0, 90000, by = 10000),
                     labels = comma)

plot_smooth_years2

# save the plot as a png file
png("plot_smooth_years2.png", width=1024, height=768, res=120)
plot(plot_smooth_years2)
dev.off()
  



# combined plot: percentage_group distribution

plot_smooth_pct2 <- ggplot(data_age_groupdistrib_BC_2019, aes(x=age_group.num, y=year_pct)) +
  geom_smooth(span = 0.3, se = FALSE, colour = "#234275") +
  geom_smooth(data = data_age_groupdistrib_BC_1971, span = 0.3, se = FALSE, colour = "#E3A82B") +
  geom_smooth(data = data_age_groupdistrib_BC_2041, span = 0.3, se = FALSE, colour = "#26BDEF") +
  annotate("text", x=20, y=2.0, label = "1971", colour =  "#234275", fontface = 2) +
  annotate("text", x=68, y=1.5, label = "2016", colour =  "#234275", fontface = 2) +
  annotate("text", x=85, y=1.0, label = "2041", colour =  "#234275", fontface = 2)
  

plot_smooth_pct2 <- plot_smooth_pct2 +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  labs(title = "British Columbia: age_group distribution (percent)",
       subtitle = "1971, 2016 & 2041",
       y = "percent of population", x = "age_group",
       caption ="Source: Statistics Canada, CANSIM Table 051-0001; BC Stats") +
  scale_x_continuous(breaks = seq(0, 100, 10)) +
  scale_y_continuous(limits = c(0, 2.25), breaks = seq(0, 2.0, by = 0.5)) 
  
plot_smooth_pct2

# save the plot as a png file
png("plot_smooth_pct2.png", width=1024, height=768, res=120)
plot(plot_smooth_pct2)
dev.off()
  

```


## Summary statistics

Questions:

- what percentage_group of British Columbians are 65 and older?
- how many school-age_group children are there?
- when was the peak in school age_group children?

-30-
