---
title: "BC_demography_01_medianage"
author: "Martin Monkman"
date: "2020-01-14"
output: html_document
---
<!--
Copyright 2020 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->


## Plot of the change in median age in British Columbia

### data

[Population estimates on July 1st, by age and sex](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710000501)
Frequency: Annual

Table: 17-10-0005-01 (formerly CANSIM 051-0001)

Geography: Canada, Province or territory


```{r, setup, echo=FALSE, message=FALSE}

# load the tidyverse packages
library(tidyverse)
library(lubridate)

# utilities
library(glue)
library(here)
library(janitor)
#

# ----
# create a chart theme & colour palette for use in ggplot2
# this theme is the basic background for a BC Stats chart
#
theme_BCStats <- 
  theme_bw() +
  theme(
    panel.border = element_rect(colour="white"),
    axis.line = element_line(colour="black"),
    legend.position=c(1,0), 
    legend.justification=c(1,0),
    legend.title = element_text(size=12),
    legend.text = element_text(size=11),
    axis.title = element_text(size=12),
    axis.text = element_text(size=12)
  )
#

# colour palette for BC Stats charts

palette_BCStats <- c("#234275", "#E3A82B", "#26BDEF", "#11CC33", "#D3E2ED", "8A8A8A")



```





```{r read_data}

# read the data
df_1710000501 <- read_rds("../data_source/df_1710000501.rds")


df_medianage <- df_1710000501 %>% 
  filter(sex == "Both sexes",
         age_group == "Median age")

# filter the long frame to include only B.C. 
df_medianage_BC <- filter(df_medianage, geo == "British Columbia")

year_min = min(df_medianage_BC$ref_date)
year_max = max(df_medianage_BC$ref_date)


```


### B.C. median age plot

The first version of the plot is a basic version of the plot, with only the line weight and colour specified.

The second version adds an assortment of formatting:

* axis length and grid specifications
* axis titles
* chart title

Note that the X axis scale is such that it follows the five year increments of the Census of Canada.





```{r, echo=FALSE, message=FALSE}

# basic
plot_median_age_BC <- 
  ggplot(df_medianage_BC, aes(x=ref_date_year, y=value)) +
  geom_line(size = 1.5)

plot_median_age_BC

# fix the formatting

plot_median_age_BC +
  theme_BCStats +
  theme(panel.grid.minor = element_blank()) +
  scale_colour_manual(values=palette_BCStats) +
  labs(title = glue('Median age, British Columbia, {year_min}–{year_max}'),
       caption = "Source: Statistics Canada. Table 17-10-0005-01",
       y = "median age",
       x = NULL) +
  scale_x_continuous(breaks = seq(year_min, year_max, by = 5)) +
  scale_y_continuous(limits = c(20, 50), breaks = seq(20, 50, by = 5)) +
  # annotations
  annotate("text", x=1971, y=22, label = "27", colour =  "#234275", fontface = 2) +
  geom_segment(aes(x=1971, xend=1971, y=23.5, yend=26.5), 
               size = 1.5, colour = "#234275",
               arrow = arrow(length = unit(.3, "cm"))) +
  annotate("text", x=2019, y=38, label = "42", colour =  "#E3A82B", fontface = 2) +
  geom_segment(aes(x=2019, xend=2019, y=38.5, yend=41.5), 
               size = 1.5, colour = "#E3A82B",
               arrow = arrow(length = unit(.3, "cm"))) 
  
  
# save the plot as a png file
ggsave("../plot_images/plot_median_age_BC.png", width=9, height=6)


```

### B.C. and Canada median age plot

This version adds a second line to allow for the comparison of the change in the median age of British Columbians to that of Canadians overall.


```{r, echo=FALSE, message=FALSE}

# filter the long frame to include only B.C. and Canada
regions <- c("British Columbia", "Canada")
df_medianage_BC_Can <- 
  filter(df_medianage, geo %in% regions)

# basic
plot_median_age_BC_Can <- 
  ggplot(df_medianage_BC_Can, aes(x=ref_date_year, y=value, colour = geo)) +
  geom_line(size = 1.5)

plot_median_age_BC_Can

# fix the formatting
plot_median_age_BC_Can +
  theme_BCStats +
  theme(panel.grid.minor = element_blank(),
        legend.title = element_blank()) +
  scale_colour_manual(values=palette_BCStats) +
  labs(title = glue('Median age, British Columbia & Canada, {year_min}–{year_max}'),
       caption = "Source: Statistics Canada. Table 17-10-0005-01",
       y = "median age",
       x = NULL) +
  scale_x_continuous(breaks = seq(1971, 2016, by = 5)) +
  scale_y_continuous(limits = c(20, 50), breaks = seq(20, 50, by = 5))

  
# save the plot as a png file
ggsave("../plot_images/plot_median_age_BC_CAN.png", width=9, height=6)

```

### Conventional wisdom: B.C. is an old province

The above chart makes it plain that British Columbians are indeed older than Canadians as a whole. 

But what may not be known is that British Columbia, while once the oldest province, no longer has that distinction. 

Between 1994 and 2001, Quebec had the highest median age. From 2002 through 2004, Nova Scotians had the highest median age. And since then, Newfoundland & Labrador has had the highest median age of any Canadian province or territory.  

The chart below compares the median age in British Columbia in the years 1971 through 2019 to Newfoundland & Labrador. It has been a remarkable change there. For the years 1971 through 1993, it was the youngest province. Over the span of a dozen years, it went from the youngest to the oldest. And in 2015 the median age in Newfoundland & Labrador was 45--a full 4.5 years higher than the median age across Canada.

This dramatic shift in the age profile in Newfoundland & Labrador was a consequence of the collapse of the North Atlantic cod fishery; the moratorium instantly put 30,000 people out of work, and the population of the province fell by 10% between 1991 and 2001. [Source: Heritage Newfoundland & Labrador, "Rural Depopulation"](http://www.heritage.nf.ca/articles/society/depopulation.php)

```{r, echo=FALSE, message=FALSE}

# filter the long frame to include B.C. and Newfoundland and Labrador
regions <- c("British Columbia", "Newfoundland and Labrador")
df_medianage_BC_prov <- 
  filter(df_medianage, geo %in% regions)

# basic
plot_median_age_BC_prov <- 
  ggplot(df_medianage_BC_prov, aes(x=ref_date_year, y=value, colour = geo)) +
  geom_line(size = 1.5)

# fix the formatting
plot_median_age_BC_prov +
  theme_BCStats +
  theme(panel.grid.minor = element_blank(),
        legend.title = element_blank()) +
  scale_colour_manual(values=palette_BCStats) +
  labs(title = glue('Median age, British Columbia and Newfoundland & Labrador, {year_min}–{year_max}'),
       caption = "Source: Statistics Canada. Table 17-10-0005-01",
       y = "median age",
       x = NULL) +
  scale_x_continuous(breaks = seq(1971, 2016, by = 5)) +
  scale_y_continuous(limits = c(20, 50), breaks = seq(20, 50, by = 5))

  
# save the plot as a png file
ggsave("../plot_images/plot_median_age_BC_NL.png", width=9, height=6)

```


-30-