---
title: "BC_demography_03_fertilityrates"
author: "Martin Monkman"
date: "February 22, 2016"
output: html_document
---
<!--
Copyright 2020 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->


## Fertility rates in British Columbia, 1921-2015


```{r, echo=FALSE, message=FALSE}

# THE PRELIMINARIES

# load the necessary packages
# data manipulation
library(tidyr)
library(dplyr)
# graphics
library(ggplot2)
library(scales)
library(RColorBrewer)
# tables
library(xtable)


# PLOTTING ESSENTIALS

# create a chart theme & colour palette for use in ggplot2
# this theme is the basic background for a BC Stats chart
#
theme_BCStats <- 
  theme_bw() +
  theme(
    panel.border = element_rect(colour="white"),
    axis.line = element_line(colour="black"),
    legend.position=c(1,0), 
    legend.justification=c(1,0),
    legend.title = element_text(size=12),
    legend.text = element_text(size=11),
    axis.title = element_text(size=12),
    axis.text = element_text(size=12)
  )
#

# colour palette for BC Stats charts (use with scale_colour_manual)
palette_BCStats <- c("#234275", "#E3A82B", "#26BDEF", "#11CC33", "#D3E2ED", "8A8A8A")

# grayscale for fill (use with scale_fill_manual)
palette_BCStats_fill <- c("#3F3F3F", "#ABABAB", "#DFDFDF", "#969696", "#838383", "8A8A8A")



```



### data

Source file:  (./data_source/20160216_AFSR.back.to.1921.xlsx)

The Excel file was cleaned up to produce a wide table that contained only the year (column 1) and age groups (columns 2-9). This table was saved as `fertilityrate.csv`.

The table is then melted into long format.



```{r, echo=FALSE, message=FALSE}

# read the data
#setwd("~/GitHub/BC_demography")
#
fertilityrate <- read.csv("./data_source/fertilityrate.csv", header = TRUE)

#colnames(data_agedistrib01) <- c("year", "All.ages", 
#                                 "00", "01", "02", "03", "04", 
#                                 "05", "06", "07", "08", "09",
#                                 seq(10, 89, by = 1), "90+")

# turn the wide version of data_medianage into long format
#data_agedistrib01_long <- data_agedistrib01 %>%
#  gather(age, popul, -year) %>%
#  filter(age != "All.ages")

#head(data_agedistrib01_long)

# turn the wide version of data_medianage into long format, add pctage column

data_fertilityrate_long <- fertilityrate %>%
  gather(age, fert_rate, -Year)

#head(data_fertilityrate_long)
#tail(data_fertilityrate_long)


```

#### Total fertility rate in B.C., 1921 - 2015

A plot showing the fertility rate in British Columbia, from 1921 through 2015.

The first version of the plot is a basic version of the plot. The second version adds an assortment of formatting:

* axis length and grid specifications
* axis titles
* chart title

Note that the X axis scale is such that it follows the five year increments of the Census of Canada.



```{r, echo=FALSE, message=FALSE}

# filter the data to include just the total

data_fertilityrate_19212015_total <- fertilityrate %>%
  gather(age, fert_rate, -Year) %>%
  filter(Year < 2016 & age == "Total")

# --------

# basic
plot_totalfert <- ggplot(data_fertilityrate_19212015_total, aes(x=Year, y=fert_rate, colour=age)) +
  #  geom_bar(stat="identity")
  geom_line(size=1.5)

# basic plot
#plot_totalfert

# fix the formatting

plot_totalfert +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  ggtitle("British Columbia, Total fertility, 1921-2015") +
  ylab("fertility rate") +
  scale_x_continuous(breaks = seq(1921, 2016, by = 5)) +
  scale_y_continuous(limits = c(0, 5000), breaks = seq(0, 5000, by = 1000),
                     labels = comma) +
  # remove the legend (not much point when there's only one line!)
  theme(legend.position="none")

```


```{r, echo=FALSE, message=FALSE}

# filter the data to include just the total

data_fertilityrate_19212015_agegrp <- fertilityrate %>%
  gather(age, fert_rate, -Year) %>%
  filter(Year < 2016 & age != "Total")

# --------

# basic
plot_fert_age <- ggplot(data_fertilityrate_19212015_agegrp, aes(x=Year, y=fert_rate, colour=age)) +
  #  geom_bar(stat="identity")
  geom_line(size=1.5)

# basic plot
plot_fert_age

# fix the formatting

plot_fert_age +
  theme_BCStats +
  scale_colour_brewer(palette="Greys") +
  ggtitle("British Columbia, Total fertility, 1921-2015") +
  ylab("fertility rate") +
  scale_x_continuous(breaks = seq(1921, 2016, by = 5)) +
  scale_y_continuous(limits = c(0, 300), breaks = seq(0, 300, by = 50),
                     labels = comma) +
  theme(legend.position=c(1,0.5))

```

Too many lines!  Let's try that as a set of facets


```{r, echo=FALSE, message=FALSE}

plot_fert_age_facet <- ggplot(data_fertilityrate_19212015_agegrp, aes(x=Year, y=fert_rate)) +
  #  geom_bar(stat="identity")
  geom_line(size=1.5) +
  facet_grid(age ~.)

# basic plot
plot_fert_age_facet

# fix the formatting

plot_fert_age_facet +
  theme_BCStats +
  scale_colour_brewer(palette="Greys") +
  ggtitle("British Columbia, Total fertility, 1921-2015") +
  ylab("fertility rate") +
  scale_x_continuous(breaks = seq(1921, 2016, by = 5)) +
  scale_y_continuous(limits = c(0, 300), breaks = seq(0, 300, by = 50),
                     labels = comma) 

```

Well that's not really doing it either.

So perhaps the first version is best, but with either the default garish colours or fewer categories. (After all, there is very low fertility in the 15 - 19 and 45 - 49 categories.)



-30-