---
title: "BC_demography_01_medianage"
author: "Martin Monkman"
date: "2020-01-14"
output: html_document
---
<!--
Copyright 2020 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->


With most of the various data files needed for these now available on CANSIM, the script below downloads those using the {cansim} package and then saves them as RDS files for later use.


```{r setup, echo=FALSE, message=FALSE}

# load the tidyverse packages
library(tidyverse)
library(lubridate)

# utilities
library(glue)
library(here)
library(janitor)
#
library(cansim)


```



## Population estimates

### Province

[Population estimates on July 1st, by age and sex](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710000501)
Frequency: Annual

Table: 17-10-0005-01 (formerly CANSIM 051-0001)

Geography: Canada, Province or territory



**Note:** the Northwest Territories splits in 1991, into Northwest Territories and Nunavut.



```{r read_data}

# read the data
df_1710000501 <- get_cansim("17-10-0005-01")

df_1710000501 <- clean_names(df_1710000501)

df_1710000501 <- df_1710000501 %>% 
  mutate(ref_date_year = as.integer(ref_date))

write_rds(df_1710000501, "../data_source/df_1710000501.rds")

```


### Census Division

Population estimates, July 1, by census division, 2016 boundaries1, 2, 3, 4, 5, 6
Frequency: Annual

Table: 17-10-0139-01

Release date: 2021-01-14

Geography: Province or territory, Census division

https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710013901



### Census Sub-Division


https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710014201

Population estimates, July 1, by census subdivision, 2016 boundaries1, 2, 3, 4, 5, 6
Frequency: Annual

Table: 17-10-0142-01

Release date: 2021-01-14

Geography: Province or territory, Census subdivision


```{r}


# read the data
df_1710014201 <- get_cansim("17-10-0142-01")

df_1710014201 <- clean_names(df_1710014201)

df_1710014201 <- df_1710014201 %>% 
  mutate(ref_date_year = as.integer(ref_date))

write_rds(df_1710014201, "../data_source/df_1710014201.rds")



```






## Population projections

Population Projections for Canada (2018 to 2068), Provinces and Territories (2018 to 2043)

https://www150.statcan.gc.ca/n1/pub/91-520-x/91-520-x2019001-eng.htm

Data: 

Projected population, by projection scenario, age and sex, as of July 1 (x 1,000)1
Frequency: Annual

Table: 17-10-0057-01 (formerly CANSIM 052-0005)

Geography: Canada, Province or territory


```{r}

# read the data
df_1710005701 <- get_cansim("17-10-0057-01")

df_1710005701 <- clean_names(df_1710005701)


write_rds(df_1710005701, "../data_source/df_1710005701.rds")



```



#### read the data


```{r read_data_projection, echo=FALSE, message_group=FALSE}

#data_BC_projection <- read_csv("./data_source/Project_Inputs_TBS_Proj1607_5_single-all.csv")


# read the data
df_1710005701 <- get_cansim("17-10-0057-01")

df_1710005701 <- clean_names(df_1710005701)


write_rds(df_1710005701, "../data_source/df_1710005701.rds")


```

Create BC-only table with the "M3" projection scenario:

```{r}


data_BC_projection <- df_1710005701 %>% 
  filter(geo == "British Columbia" &
           projection_scenario == "Projection scenario M3: medium-growth")


write_rds(data_BC_projection, "../data_source/data_BC_projection.rds")


```


-30-