---
title: "BC demography"
author: "Martin Monkman"
date: "2021-11-19"
output: html_document
---
<!--
Copyright 2020 & 2021 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->


## British Columbia Population: Census Subdivisions

### data

Population estimates, July 1, by census subdivision, 2016 boundaries1, 2, 3, 4, 5, 6
Frequency: Annual

Table: 17-10-0142-01

Release date: 2021-01-14

Geography: Province or territory, Census subdivision

https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710014201


```{r setup, echo=FALSE, message=FALSE}

# load the tidyverse packages
library(tidyverse)
library(lubridate)

# utilities
library(glue)
library(here)
library(janitor)
#

# ----
# create a chart theme & colour palette for use in ggplot2
# this theme is the basic background for a BC Stats chart
#
theme_BCStats <- 
  theme_bw() +
  theme(
    panel.border = element_rect(colour="white"),
    axis.line = element_line(colour="black"),
    legend.position=c(1,0), 
    legend.justification=c(1,0),
    legend.title = element_text(size=12),
    legend.text = element_text(size=11),
    axis.title = element_text(size=12),
    axis.text = element_text(size=12)
  )
#

# colour palette for BC Stats charts

palette_BCStats <- c("#234275", "#E3A82B", "#26BDEF", "#11CC33", "#D3E2ED", "8A8A8A")


# ---
# CAGR funtion to calculate annual average growth


fun_cagr <- function(datavar, n) {
  mutate(cagr = ((datavar / lag(datavar, n)) ^ (1 / n)) - 1)
  return(datavar)
}

```





```{r read_data}

# read the data
df_1710014201 <- read_rds("../data_source/df_1710014201.rds")

ls.str(df_1710014201)

```



```{r data_filter_bc}
# filter for British Columbia

df_bc_pop <- df_1710014201 %>% 
  filter(str_detect(geo, "British Columbia")) %>% 
  mutate(cd = str_sub(dguid, start = 10, end = 13), 
         csd = str_sub(dguid, start = 12, end = 16))

df_bc_pop

#write_rds(df_bc_pop, "../data_source/df_1710014201_bc_csd.rds")

df_bc_pop <- df_bc_pop %>% 
  select(ref_date_year, geo, cd, csd, value)

df_bc_pop

```

quick look--are the csd values unique to the geo?

```{r}
df_bc_pop %>% 
  filter(ref_date_year == 2020,
         cd == "5917") %>% 
  arrange(csd)
```





## Boundaries Commission

### BC total

```{r cagr_bc}

# Electoral Boundary Commission years
ebc_years <- c(2008, 2015, 2020)

df_bc_pop %>%
  # filter for geography & relevant years (previous commissions)
  filter(geo == "British Columbia") %>%
  filter(ref_date_year %in% ebc_years) %>%
  #
  select(geo, ref_date_year, value) %>%
  # total growth between commissions
  mutate(pop_change = value - lag(value, 1),
         pop_change_pct = round((pop_change / lag(value, 1) * 100), 2)) %>%
# compound annual growth rate (CAGR) between commissions
  mutate(cagr_years = ref_date_year - lag(ref_date_year, 1)) %>%
  mutate(cagr = (((value / lag(value, 1)) ^ (1 / cagr_years) - 1) * 100))

```


### CSD

```{r}

df_bc_pop2 <- df_bc_pop %>% 
  # filter out summary rows
  filter(geo != "British Columbia") %>% 
  filter(geo != "Census subdivisions with 5,000-plus population, British Columbia") %>% 
  filter(geo != "Census subdivisions with population under 5,000, British Columbia") %>% 
  # filter for relevant years (previous commissions)
  filter(ref_date_year %in% ebc_years) %>% 
  arrange(csd) 

df_bc_pop2

```


```{r}
df_bc_pop2 <- df_bc_pop2 %>% 
  group_by(geo) %>% 
  # total growth between commissions
  mutate(pop_change = value - lag(value, 1),
         pop_change_pct = round((pop_change / lag(value, 1) * 100), 2)) %>%
# compound annual growth rate (CAGR) between commissions
  mutate(cagr_years = ref_date_year - lag(ref_date_year, 1)) %>%
  mutate(cagr = (((value / lag(value, 1)) ^ (1 / cagr_years) - 1) * 100))

df_bc_pop2
```

make table where population in 2020 is in excess of 10000


```{r}
df_pop_10k <- df_bc_pop2 %>% 
  filter(ref_date_year == 2020) %>% 
  filter(value >= 10000) %>% 
  arrange(csd)

df_pop_10k

```

use semi-join to filter main table (so only those CSDs with 2020 pop > 10000 remain)

```{r}

df_bc_pop3 <- df_bc_pop2 %>% 
  semi_join(df_pop_10k, by = "geo")


df_bc_pop3

```


### CD

```{r}
df_bc_pop2 %>% 
  mutate(cagr_years = ref_date_year - lag(ref_date_year, 1))

df_bc_pop2 %>% 
  group_by(cd, ref_date_year) %>% 
  summarise(value = sum(value)) %>% 
  #
  group_by(cd) %>% 
  # total growth between commissions
  mutate(pop_change = value - lag(value, 1),
         pop_change_pct = round((pop_change / lag(value, 1) * 100), 2)) %>%
# compound annual growth rate (CAGR) between commissions
  mutate(cagr_years = ref_date_year - lag(ref_date_year, 1)) %>%
  mutate(cagr = (((value / lag(value, 1)) ^ (1 / cagr_years) - 1) * 100))
  
```




-30-