---
title: "BC_demography_04b_mortality"
author: "Martin Monkman"
date: "February 23, 2016"
output: html_document
---
<!--
Copyright 2020 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->


## Mortality rates in British Columbia, 1991-2012


**Note:** this version calculates the mortality rates from scratch, allowing the period to extend back to 1971 and forward to 2015 (compared to the published Statisics Canada table, which only covers 1991-2012).



```{r, echo=FALSE, message=FALSE}

# THE PRELIMINARIES

# load the necessary packages
# data manipulation
library(tidyr)
library(dplyr)
# graphics
library(ggplot2)
library(scales)
library(RColorBrewer)
# tables
library(xtable)


# PLOTTING ESSENTIALS

# create a chart theme & colour palette for use in ggplot2
# this theme is the basic background for a BC Stats chart
#
theme_BCStats <- 
  theme_bw() +
  theme(
    panel.border = element_rect(colour="white"),
    axis.line = element_line(colour="black"),
    legend.position=c(1,0), 
    legend.justification=c(1,0),
    legend.title = element_text(size=12),
    legend.text = element_text(size=11),
    axis.title = element_text(size=12),
    axis.text = element_text(size=12)
  )
#

# colour palette for BC Stats charts (use with scale_colour_manual)
palette_BCStats <- c("#234275", "#E3A82B", "#26BDEF", "#11CC33", "#D3E2ED", "8A8A8A")

# grayscale for fill (use with scale_fill_manual)
palette_BCStats_fill <- c("#3F3F3F", "#ABABAB", "#DFDFDF", "#969696", "#838383", "8A8A8A")

# NOTE: the RColorBrewer palette "Greys" was used in order to provide more shades


```


The calculation of mortality rates is in two stages:

1. Ages 1 to 90+

*  group into five year age groups (starting with 1-4 [not 5 years I know!], then 5-9, 10-14, etc.)
*  sum the deaths and the population in those groups
*  (numerator) deaths in age category divided by (denominator) total population, multiplied by 1,000

2. Under age 1 (infant mortality)


### data


#### population

Source file:  `AS_1971_2015_v1.xlsx`

* the population sheet was saved as a separate XLSX file, transformed, and a ReadMe sheet spelling out the details of the transformation.

* The transformed table was saved as `BC_DEATHS_1971-2015.csv`.

The table is read into the workspace, then melted into long format.



```{r, echo=FALSE, message=FALSE}

# read the data
#setwd("~/GitHub/BC_demography")
#

BC_population <- read.csv("./data_source/BC_Population_1971-2015.csv", header = TRUE)

# SAVE FOR LATER!
#BC_deaths <- read.csv("./data_source/BC_DEATHS_1971-2015.csv", header = TRUE)

#head(mortalityrate)

# clean up the variable names
colnames(BC_population) <- c("SGC", "Area.Name", "Year", "Level", "Sex",
                             "Total", "under1", "00", "01", "02", "03", "04", 
                             "05", "06", "07", "08", "09", 
                             seq(10, 89, by = 1), "90+")

head(BC_population)


# convert to long format

BC_population_long <- BC_population %>%
  select(-(SGC:Area.Name), -Level) %>%
  gather(Age, Popul, -c(Sex, Year))

head(BC_population_long)


# define the five year groups

Infant <- c("under1", "00")
age01_04 <- c("01", "02", "03", "04")
age05_09 <- c("05", "06", "07", "08", "09")
age10_14 <- c(seq(10, 14, by = 1))
age15_19 <- c(seq(15, 19, by = 1))
age20_24 <- c(seq(20, 24, by = 1))
age25_29 <- c(seq(25, 29, by = 1))
age30_34 <- c(seq(30, 34, by = 1))
age35_39 <- c(seq(35, 39, by = 1))
age40_44 <- c(seq(40, 44, by = 1))
age45_49 <- c(seq(45, 49, by = 1))
age50_54 <- c(seq(50, 54, by = 1))
age55_59 <- c(seq(55, 59, by = 1))
age60_64 <- c(seq(60, 64, by = 1))
age65_69 <- c(seq(65, 69, by = 1))
age70_74 <- c(seq(70, 74, by = 1))
age75_79 <- c(seq(75, 79, by = 1))
age80_84 <- c(seq(80, 84, by = 1))
age85_89 <- c(seq(85, 89, by = 1))
age90plus <- c("90+")


BC_population_long$AgeGrp[BC_population_long$Age == "Total"] <- "Total"
BC_population_long$AgeGrp[BC_population_long$Age %in% Infant] <- "Infant"
BC_population_long$AgeGrp[BC_population_long$Age %in% age01_04] <- "01_04"
BC_population_long$AgeGrp[BC_population_long$Age %in% age05_09] <- "05_09"
BC_population_long$AgeGrp[BC_population_long$Age %in% age10_14] <- "10_14"
BC_population_long$AgeGrp[BC_population_long$Age %in% age15_19] <- "15_19"
BC_population_long$AgeGrp[BC_population_long$Age %in% age20_24] <- "20_24"
BC_population_long$AgeGrp[BC_population_long$Age %in% age25_29] <- "25_29"
BC_population_long$AgeGrp[BC_population_long$Age %in% age30_34] <- "30_34"
BC_population_long$AgeGrp[BC_population_long$Age %in% age35_39] <- "35_39"
BC_population_long$AgeGrp[BC_population_long$Age %in% age40_44] <- "40_44"
BC_population_long$AgeGrp[BC_population_long$Age %in% age45_49] <- "45_49"
BC_population_long$AgeGrp[BC_population_long$Age %in% age50_54] <- "50_54"
BC_population_long$AgeGrp[BC_population_long$Age %in% age55_59] <- "55_59"
BC_population_long$AgeGrp[BC_population_long$Age %in% age60_64] <- "60_64"
BC_population_long$AgeGrp[BC_population_long$Age %in% age65_69] <- "65_69"
BC_population_long$AgeGrp[BC_population_long$Age %in% age70_74] <- "70_74"
BC_population_long$AgeGrp[BC_population_long$Age %in% age75_79] <- "75_79"
BC_population_long$AgeGrp[BC_population_long$Age %in% age80_84] <- "80_84"
BC_population_long$AgeGrp[BC_population_long$Age %in% age85_89] <- "85_89"
BC_population_long$AgeGrp[BC_population_long$Age == "90+"] <- "90+"





head(BC_population_long)


# ------------ START HERE -------------------

# sum the five-year groups

BC_population_grps <- BC_population_long %>%
  # filter in the Sex == 3 (Both), filter out the under 1 and 0 ages
  filter(Sex == "3", !(Age %in% c("Total", "under1", "00"))) %>%
  # summarize into the age categories


head(BC_population_grps)
















mortalityrate$Age1[mortalityrate$Age == "all ages"] <- "all ages"
mortalityrate$Age1[mortalityrate$Age == "under 1 year"] <- "00"
mortalityrate$Age1[mortalityrate$Age == "1 to 4 years"] <- "01-04"
mortalityrate$Age1[mortalityrate$Age == "5 to 9 years"] <- "05-09"
mortalityrate$Age1[mortalityrate$Age == "10 to 14 years"] <- "10-14"
mortalityrate$Age1[mortalityrate$Age == "15 to 19 years"] <- "15-19"
mortalityrate$Age1[mortalityrate$Age == "20 to 24 years"] <- "20-24"
mortalityrate$Age1[mortalityrate$Age == "25 to 29 years"] <- "25-29"
mortalityrate$Age1[mortalityrate$Age == "30 to 34 years"] <- "30-34"
mortalityrate$Age1[mortalityrate$Age == "35 to 39 years"] <- "35-39"
mortalityrate$Age1[mortalityrate$Age == "40 to 44 years"] <- "40-44"
mortalityrate$Age1[mortalityrate$Age == "45 to 49 years"] <- "45-49"
mortalityrate$Age1[mortalityrate$Age == "50 to 54 years"] <- "50-54"
mortalityrate$Age1[mortalityrate$Age == "55 to 59 years"] <- "55-59"
mortalityrate$Age1[mortalityrate$Age == "60 to 64 years"] <- "60-64"
mortalityrate$Age1[mortalityrate$Age == "65 to 69 years"] <- "65-69"
mortalityrate$Age1[mortalityrate$Age == "70 to 74 years"] <- "70-74"
mortalityrate$Age1[mortalityrate$Age == "75 to 79 years"] <- "75-79"
mortalityrate$Age1[mortalityrate$Age == "80 to 84 years"] <- "80-84"
mortalityrate$Age1[mortalityrate$Age == "85 to 89 years"] <- "85-89"
mortalityrate$Age1[mortalityrate$Age == "90 years and over"] <- "90+"

#head(mortalityrate)
#tail(mortalityrate)


# turn the wide version of data_medianage into long format

data_mortalityrate_long <- mortalityrate %>%
  select(-Age) %>%
  gather(year, mort_rate, -Age1, -Sex)

#head(data_mortalityrate_long)
#tail(data_mortalityrate_long)

```

#### Total mortality rate in B.C., 1991 - 2012

A plot showing the mortality rate in British Columbia, from 1991 through 2012.

The first plot shows the total mortality rate, which is little changed over the period.

Note that the X axis scale is such that it follows the five year increments of the Census of Canada.



```{r, echo=FALSE, message=FALSE}

# filter the data to include just the total

data_mortalityrate_total <- mortalityrate %>%
  select(-Age) %>%
  gather(year, mort_rate, -Age1, -Sex) %>%
  filter(Sex == "Both sexes", Age1 == "all ages")

#head(data_mortalityrate_total)
#ls.str(data_mortalityrate_total)

# ------

# basic
plot_totalmort <- ggplot(data_mortalityrate_total, aes(x=year, y=mort_rate, group=Age1)) +
  #  geom_bar(stat="identity")
  geom_line(size=1.5)

# basic plot
#plot_totalmort

# fix the formatting

plot_totalmort +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  ggtitle("British Columbia, Total mortality, 1991-2012") +
  ylab("mortality rate") +
  scale_x_discrete(breaks = seq(1991, 2012, by = 5)) +
  scale_y_continuous(limits = c(6, 8), breaks = seq(6, 8, by = 0.5)) +
  # remove the legend (not much point when there's only one line!)
  theme(legend.position="none")

```


We begin to see the improvements in mortality if we look across the age ranges (along the bottom axis). Even at this scale it is possible to see that fewer infants are dying before their first birthday. And the gap begins to widen starting at the 60-64 age group.


```{r, echo=FALSE, message=FALSE}

# filter to include the age categories


data_mortalityrate_annual <- mortalityrate %>%
  select(-Age) %>%
  gather(year, mort_rate, -Age1, -Sex) %>%
  filter(Sex == "Both sexes", Age1 != "all ages")

#head(data_mortalityrate_annual)
#ls.str(data_mortalityrate_annual)


year_list <- c(1991, 2012)

data_mortalityrate_91_12 <- data_mortalityrate_annual %>%
  filter(year %in% year_list) 

#head(data_mortalityrate_91_12)
#tail(data_mortalityrate_91_12)


# --------


# basic
plot_mort_91_12 <- 
  ggplot(data_mortalityrate_91_12, aes(x=Age1, y=mort_rate, group=year, colour=year)) +
  geom_line(size=1.5)

# basic plot
#plot_mort_91_12

# fix the formatting

plot_mort_91_12 +
  theme_BCStats +
  scale_colour_manual(values=palette_BCStats) +
  ggtitle("British Columbia, Total mortality, 1991-2012") +
  ylab("mortality rate") +
#  scale_x_discrete(breaks = seq(1991, 2012, by = 5)) +
  scale_y_continuous(limits = c(0, 225), breaks = seq(0, 200, by = 50)) +
  # put the legend in the top left corner
  theme(legend.position = c(0.15, 0.75))


```

**Note**: running into William Playfair's balance of trade visualization problem here. Would be better to show the improvement.


-30-