---
title: "Canada_demography_02_agedistribution"
author: "Martin Monkman"
date: "2021-06-11"
output: html_document
---
<!--
Copyright 2020 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->


```{r setup, echo=FALSE, message_group=FALSE}

# tidyverse
library(tidyverse)

# utilities
library(cansim)
library(glue)
library(here)
library(janitor)
#
# tables
library(xtable)
#
# plotting
library(gganimate)
library(zoo)
library(scales)
#

# PLOTTING ESSENTIALS

# create a chart theme & colour palette for use in ggplot2
# this theme is the basic background for a BC Stats chart
#
theme_BCStats <- 
  theme_bw() +
  theme(
    panel.border = element_rect(colour="white"),
    axis.line = element_line(colour="black"),
#    legend.position=c(1,0), 
#    legend.justification=c(1,0),
    legend.title = element_text(size=12),
    legend.text = element_text(size=11),
    axis.title = element_text(size=16),
    axis.text = element_text(size=16),
    plot.title = element_text(size=18)
  )
#

# colour palette for BC Stats charts (use with scale_colour_manual)
palette_BCStats <- c("#234275", "#E3A82B", "#26BDEF", "#11CC33", "#D3E2ED", "8A8A8A")

# grayscale for fill (use with scale_fill_manual)
palette_BCStats_fill <- c("#3F3F3F", "#ABABAB", "#DFDFDF", "#969696", "#838383", "8A8A8A")


#
txt_caption <- "Source: Statistics Canada, CANSIM Table 17-10-0005-01"
txt_caption2 <- "Source: Statistics Canada, CANSIM Tables 17-10-0005-01 & 17-10-0057-01"


```

## Plot the age distribution of British Columbia & Canada, 1971–present


### data: estimates

[Population estimates on July 1st, by age and sex](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710000501&pickMembers%5B0%5D=1.1&pickMembers%5B1%5D=2.1&cubeTimeFrame.startYear=2016&cubeTimeFrame.endYear=2020&referencePeriods=20160101%2C20200101)
Frequency: Annual

Table: 17-10-0005-01 (formerly CANSIM 051-0001)

Geography: Canada, Province or territory


### data: projections


Source:
Statistics Canada, CANSIM 
[Projected population, by projection scenario, age and sex, as of July 1 (x 1,000)](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710005701)
Frequency: Annual

Table: 17-10-0057-01 (formerly CANSIM 052-0005)


Read and clean data

```{r read_data, echo=FALSE, message_group=FALSE}

# read the data
df_1710005701 <- get_cansim("17-10-0057-01")

df_1710005701 <- clean_names(df_1710005701)

df_1710005701_med <- df_1710005701 %>% 
  filter(projection_scenario == "Projection scenario M1: medium-growth",
         sex == "Both sexes",
         geo == "Canada") %>% 
  mutate(age_group = as_factor(age_group))

```





Filter so that only single year categories remain (i.e. age 0, 1, 2, etc)

**Note:** the years up to and including 2000 include "90 years and over" as the oldest category; in 2001 the years 90-99 were included as single years, and the oldest category became "100 years and over". To make a data frame with single categories

* for those years, the 90+ category needs to be reframed as the single year "90"

* for the other years, the 90+ category needs to be dropped entirely

```{r}
# age category values to drop
drop.age_group.vals_1 <- c("All ages", "0 to 4 years", "5 to 9 years",
                   "10 to 14 years", "15 to 19 years", 
                   "20 to 24 years", "25 to 29 years", 
                   "30 to 34 years", "35 to 39 years", 
                   "40 to 44 years", "45 to 49 years", 
                   "50 to 54 years", "55 to 59 years", 
                   "60 to 64 years", "65 to 69 years", 
                   "70 to 74 years", "75 to 79 years", 
                   "80 to 84 years", "85 to 89 years", 
                   "90 to 94 years", "95 to 99 years",
                   "0 to 14 years", "0 to 15 years", "0 to 16 years", "0 to 17 years",
                   "15 to 49 years", "15 to 64 years", "16 to 64 years", "17 to 64 years",
                   "18 years and over", "18 to 64 years", "18 to 24 years", "25 to 44 years",
                   "45 to 64 years", "65 years and over", 
                   "Median age", "Average age")

df_single <- df_1710005701_med %>% 
  filter(!age_group %in% drop.age_group.vals_1)


#write_rds(df_single, "../data_output/Canada__proj_age_single.rds")
```



```{r}
# age category values to keep
age_group.vals_1 <- c("0 to 4 years", "5 to 9 years",
                   "10 to 14 years", "15 to 19 years", 
                   "20 to 24 years", "25 to 29 years", 
                   "30 to 34 years", "35 to 39 years", 
                   "40 to 44 years", "45 to 49 years", 
                   "50 to 54 years", "55 to 59 years", 
                   "60 to 64 years", "65 to 69 years", 
                   "70 to 74 years", "75 to 79 years", 
                   "80 to 84 years", "85 to 89 years", 
                   "90 to 94 years", "95 to 99 years",
                   "100 years and over")

age_group.young <- c("0 to 4 years", "5 to 9 years",
                   "10 to 14 years")

age_group.old <- c("65 to 69 years", 
                   "70 to 74 years", "75 to 79 years", 
                   "80 to 84 years", "85 to 89 years", 
                   "90 to 94 years", "95 to 99 years",
                   "100 years and over")


df_5yr_grp <- df_1710005701_med %>% 
  filter(age_group %in% age_group.vals_1)

df_5yr_grp <- df_5yr_grp %>% 
  mutate(age_grp_3 = case_when(
    age_group %in% age_group.young ~ "0-14",
    age_group %in% age_group.old ~ "65+",
    TRUE ~ NA_character_
  ))

df_5yr_grp %>% 
  filter(ref_date == 2020) %>% 
  group_by(ref_date, age_group, age_grp_3) %>% 
  summarise(population = sum(`value`)) 

```


```{r}

df_5yr_grp %>% 
  filter(ref_date == 2020) %>% 
  group_by(age_group) %>% 
  summarise(population = sum(`value`)) %>% 
  mutate(pct_pop = population / sum(population)) %>% 
  summarise(sum(population),
            sum(pct_pop))


df_5yr_grp %>% 
  filter(ref_date == 2020) %>% 
  group_by(age_group) %>% 
  summarise(population = sum(`value`)) %>% 
  mutate(pct_pop = population / sum(population)) 

df_5yr_grp %>% 
  filter(ref_date == 2020) %>% 
  group_by(age_grp_3) %>% 
  summarise(population = sum(`value`)) %>% 
  mutate(pct_pop = population / sum(population)) 


#write_rds(df_single, "../data_output/Canada__proj_age_single.rds")
```


-30-
